---
title: "Getting Started with eDNAstructure"
subtitle: "Fitting Dirichlet-Multinomial Mixture Models to eDNA Metabarcoding Data"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 3
    number_sections: true
vignette: >
  %\VignetteIndexEntry{Getting Started with eDNAstructure}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse  = TRUE,
  comment   = "#>",
  eval      = FALSE,   # set to TRUE once package is installed with rstan
  fig.width = 7,
  fig.height = 5
)
```

# Introduction

`eDNAstructure` fits **Dirichlet-Multinomial Mixture (DMM) models** to
environmental DNA (eDNA) read count data from metabarcoding surveys. The
central question the model answers is:

> *Given read count data from many sampling locations, how many distinct
> ecological communities exist in my study area; and what environmental
> variables predict which community a sample belongs to?*

The model is appropriate whenever you have:

- A matrix of read counts (samples × taxa or samples × ASVs)
- Two or more sampling locations
- An interest in whether samples cluster into distinct assemblage types
- Optionally: environmental covariates you believe drive that clustering

This vignette walks you through the complete workflow from raw data to
publication-ready figures. All code uses the built-in example dataset so
you can run everything yourself.

---

# Installation

## Step 1: Install R

If you do not yet have R installed, download the latest version for your
operating system from <https://cran.r-project.org/>. Version 4.1.0 or later
is required.

We also strongly recommend installing **RStudio**, a free graphical interface
for R: <https://posit.co/download/rstudio-desktop/>.

## Step 2: Install Stan and rstan

`eDNAstructure` uses the Stan probabilistic programming language for Bayesian
inference. Stan has its own installation requirements beyond a standard R
package (and it is usually very annoying). Follow the official instructions carefully:

**<https://mc-stan.org/rstan/articles/rstan.html>**

The key steps are:

1. Install a C++ toolchain for your OS:
   - **Windows**: Install Rtools from <https://cran.r-project.org/bin/windows/Rtools/>
   - **macOS**: Install the Xcode Command Line Tools (`xcode-select --install` in Terminal)
   - **Linux**: Ensure `build-essential` (or equivalent) is installed

2. Install `rstan` from CRAN:
   ```r
   install.packages("rstan")
   ```

3. Verify the installation works:
   ```r
   library(rstan)
   example(stan_model, package = "rstan", run.dontrun = TRUE)
   ```
   If this runs without errors and produces output like `"SAMPLING FOR MODEL..."`,
   Stan is working correctly.

## Step 3: Install eDNAstructure

Install directly from GitHub using the `remotes` package:

```r
# Install remotes if you don't have it
install.packages("remotes")

# Install eDNAstructure
remotes::install_github("pedrobdfp/eDNA_structure")
```

This will also install the required R package dependencies automatically:
`ggplot2`, `dplyr`, `tidyr`, `vegan`, `posterior`, `loo`, and `scales`.

## Step 4: Load the package

```{r load}
library(eDNAstructure)
```

---

# Your data and what the model expects

Before fitting a model, it helps to understand exactly what format the
functions expect. We will use the built-in example dataset to illustrate.
The model needs **two things**, and only the first is required:

1. **A species/ASV × site table**: the count matrix.
2. **A covariate table**: *optional*. Supply it only if you want to model how
   environmental variables drive community membership. `eDNA_dmm()` runs
   perfectly well without one.

## Starting from one file

In practice your data usually arrive as a single spreadsheet with sample IDs,
environmental variables and counts all side by side. A file of exactly that
shape ships with the package, so the split is easy to see:

```{r raw_file}
path <- system.file("extdata", "example_edna_raw.csv", package = "eDNAstructure")
raw  <- read.csv(path, check.names = FALSE)

raw[1:3, 1:7]
```

Split it into the two tables the model wants:

```{r raw_split}
# Which columns hold taxa? Adjust this one line to match your own file --
# a name prefix, a known column range, or setdiff() against your metadata names.
taxon_cols <- grep("^Sp_", names(raw))

# 1. THE SPECIES/ASV x SITE TABLE -- required
counts <- as.matrix(raw[, taxon_cols])
rownames(counts) <- raw$sample_id
storage.mode(counts) <- "integer"

# 2. THE COVARIATE TABLE -- optional, numeric columns only
covariates <- raw[, c("Depth", "Distance_shore")]
rownames(covariates) <- raw$sample_id

dim(counts)
dim(covariates)
```

Both fits are valid:

```{r raw_split_fit, eval=FALSE}
fit <- eDNA_dmm(counts, K = 4)                           # without covariates
fit <- eDNA_dmm(counts, covariates = covariates, K = 4)  # with them
```

> **The covariate table must be numeric, and only numeric.** `eDNA_dmm()` fits
> every column you hand it, so an identifier such as `sample_id`, or a label
> such as `TrueCommunity`, must not be in it, the model would try to fit the
> ID as an environmental gradient. Keep those in a separate metadata frame and
> pass it to the plotting functions' `metadata` argument instead. This is why
> `get_example_data()` returns `covariates` and `metadata` as two objects.

below.


## The count matrix

The most important input is a **sample × taxon count matrix**. The function
`eDNA_dmm()` accepts:

- A `matrix` with non-negative integer values
- Or a `data.frame` where every column is numeric

Rows represent **samples** (individual water samples or stations).
, taxonomic annotation is not required. The model is equally applicable to
raw ASV tables and to taxonomically-collapsed count tables.

```{r count_matrix}
# Load the example dataset
data <- get_example_data()

# Inspect the count matrix
dim(data$counts)  # 20 samples x 33 taxa

# First 5 samples, first 6 taxa
data$counts[1:5, 1:6]
```

The output will look something like:
```
        Sp_1 Sp_2 Sp_10 Sp_12 Sp_13 Sp_27
STN_001   34  154  3922  2552  1051   738
STN_002 1280  189  1366     0   526  1053
STN_003 1665  878   606   687   712  1018
```

Each number is the count of sequencing reads assigned to that taxon in that
sample. These are raw counts, **do not** normalize to proportions before
fitting. The model handles composition estimation internally.

**Key rules for the count matrix:**

| Requirement | What happens if violated |
|-------------|--------------------------|
| All values ≥ 0 | Error with informative message |
| No NA values | Error with informative message |
| No all-zero rows | Error: empty samples cannot be modelled |
| At least 2 columns | Error |
| Integer values | Warning, then automatic coercion |
| Row sums ≥ 100 | Warning (very low counts = unreliable compositions) |

## The covariate data frame

The second input (optional) is a **sample × covariate** matrix or data frame.
If you pass covariates, the model will estimate how each covariate shifts the
probability that a sample belongs to each community.

```{r covariates}
# Inspect the covariate data frame
head(data$covariates)   # numeric only
```

```
  SampleID TrueCommunity    Depth Distance_shore sample_id
1        1             1 86.25497       201.9523   STN_001
2        2             1 73.32899       197.4781   STN_002
3        3             1 75.75906       208.2311   STN_003
```

**Critical:** the covariate data frame must have exactly the same number of
rows as your count matrix, **in the same order**. Each row `i` of the
covariate data frame corresponds to row `i` of the count matrix. If these
are misaligned, the model will silently use the wrong covariate values for
each sample.

A safe way to ensure alignment is to match on sample IDs:

```{r align}
# Ensure both objects are sorted by sample ID before fitting
counts_sorted     <- data$counts[order(rownames(data$counts)), ]
metadata_sorted   <- data$metadata[order(data$metadata$sample_id), ]

# Verify alignment
stopifnot(all(rownames(counts_sorted) == covariates_sorted$sample_id))
```

### Which covariates to include?

Include any continuous environmental variable you hypothesize might predict
community membership:

- **Spatial**: latitude, longitude, distance from shore, distance from canyon
- **Physical**: depth, temperature, salinity, oxygen, turbidity
- **Derived**: distance from thermocline, distance to seafloor, upwelling index

Covariates are Z-score standardized internally by default (`scale_covariates = TRUE`).
This means the fitted beta coefficients are directly comparable across
covariates regardless of their original units. Do not standardize your
covariates yourself unless you set `scale_covariates = FALSE`.

Categorical variables (e.g., site, year) can be included as numeric dummy
indicators, but interpret with care: the model treats them as continuous.
For categorical grouping it is often better to use them for faceting in
plots (post-hoc) rather than as covariates.

### Preparing a long-format table

If your data come in long format (one row per sample-taxon combination),
convert to wide format before using this package:

```{r long_to_wide, eval=FALSE}
library(tidyr)
library(dplyr)

# Assume `long_df` has columns: sample_id, taxon, reads
count_matrix <- long_df %>%
  pivot_wider(
    names_from  = taxon,
    values_from = reads,
    values_fill = 0          # replace missing with 0
  ) %>%
  tibble::column_to_rownames("sample_id") %>%
  as.matrix()
```

---

# Fitting the model

## Basic fit: K = 2

Start simple. A K = 2 model answers: "can my samples be divided into two
distinct communities?"

```{r fit_k2}
# Select which covariates to use
covariate_matrix <- data$covariates   # already numeric-only

# Fit a K=2 DMM
fit_k2 <- eDNA_dmm(
  counts     = data$counts,
  covariates = covariate_matrix,
  K          = 2
)
```

When you run this, you will see Stan's sampling progress printed to the
console. This is normal. Four chains of 4000 iterations (2000 warmup)
typically takes 1–5 minutes depending on N and S.

```r
## Fitting DMM: N=20 samples, S=33 taxa, K=2 communities, P=2 covariates
## MCMC: 1 chain(s), 4000 iterations (2000 warmup, 2000 sampling)
## ...
## Chain 1: Iteration: 1 / 4000 [  0%]  (Warmup)
## Chain 1: Iteration: 400 / 4000 [ 10%]  (Warmup)
## ...
## Chain 1: Sampling completed.
```

### Why one chain?

Mixture models have a fundamental ambiguity: "Community 1" in one chain
might correspond to "Community 2" in another (label switching). When chains
are combined, standard convergence diagnostics (Rhat) appear terrible even
when each individual chain has converged perfectly. One long chain avoids
this completely. The within-chain ESS reported in `summary()` is the relevant
convergence diagnostic.

## Examining the results

```{r print_fit}
# Quick overview
print(fit_k2)
```

```
eDNA Dirichlet-Multinomial Mixture Model
=========================================
  K (communities)  : 2
  N (samples)      : 60
  S (taxa)         : 25
  Covariates       : Depth, Distance_shore
  Mean alpha       : 2.84

Community sizes (MAP assignment):
  Community 1     : 32 samples (53.3%)
  Community 2     : 28 samples (46.7%)
```

```{r summary_fit}
# Full diagnostics
summary(fit_k2)
```

The summary prints:

1. **Alpha (overdispersion)**: typical values for eDNA data are 1–5.
   High alpha (>10) means counts are nearly multinomial (low overdispersion).
   Low alpha (~1) means very high overdispersion, typical for eDNA.

2. **Assignment certainty**: the fraction of samples with ≥80% probability
   of belonging to one community. High certainty (>90%) means the two
   communities are well-separated.

3. **Beta coefficients**: the effect of each covariate on community membership
   (relative to the reference community). A 90% CI that does not include 0 is
   evidence for a real effect.

4. **HMC diagnostics**: look for "No divergences", "No max treedepth",
   "No low BFMI". Any warnings here should be addressed before interpreting
   results.

## Accessing model components directly

The fit object is a standard R list. You can access all components directly:

```{r access}
# Posterior community membership probabilities (one row per sample)
head(fit_k2$sample_info)
# Columns: sample_id, prob_comm1, prob_comm2, z_hat, assignment_certainty

# Posterior mean community compositions (K x S matrix)
fit_k2$pi_mean          # rows = communities, columns = taxa
round(fit_k2$pi_mean[, 1:5], 3)

# Beta coefficient posterior summaries
fit_k2$beta_summary

# Posterior mean alpha
fit_k2$alpha_mean
```


# Choosing K: How many communities?

In practice, you rarely know K in advance. The recommended approach is to fit
models for K = 2 through K = 6 (or however many is ecologically plausible)
and compare them using Leave-One-Out cross-validation (LOO-CV).

## Using eDNA_loo()

```{r loo}
# Fit K=2 through K=5 and compare (this takes several minutes)
loo_result <- eDNA_loo(
  counts     = data$counts,
  covariates = covariate_matrix,
  K_range    = 2:5,
  iter       = 4000,
  warmup     = 2000
)
```

```{r loo_plot}
# Elbow plot
loo_result$plot
```

```{r loo_table}
# LOO-ELPD table
loo_result$loo_table
```

```
  K       elpd        se
1 2 -12043.21  81.34
2 3 -11812.55  78.92
3 4 -11834.77  79.61
4 5 -11849.03  80.15
```

**Reading the elbow plot:**

- **Higher LOO-ELPD** = better predictive performance = better model
- The "elbow" is the K beyond which improvements are negligible
- If two models are within ~2 SE of each other, prefer the simpler one (lower K)
- Biologically, K should also be interpretable, resist the urge to over-split

For the example data (which was generated with K = 3), the LOO-ELPD should
peak around K = 3 before levelling off. The model successfully recovers the
true number of communities.

## Fitting the selected K

Once you have chosen K, fit the final model with more iterations for better
posterior summaries:

```{r fit_final}
fit <- eDNA_dmm(
  counts     = data$counts,
  covariates = covariate_matrix,
  K          = 3,
  iter       = 4000,
  warmup     = 2000
)
```

---

# Visualization

All four visualization functions return `ggplot2` objects. This means you
can:

1. Print them directly to see the plot
2. Save them with `ggsave()`
3. Modify them by adding ggplot2 layers
4. Combine them with `cowplot::plot_grid()` or `patchwork`

> **One exception.** `eDNA_dmm_structure()` and `plot_true_compositions()`
> return a *cowplot grid* object rather than a ggplot when `facet_row_var` is
> supplied, because they build one sub-plot per row level to keep panel widths
> proportional to sample count. Adding ggplot2 layers with `+` will not work on
> those; style them through the functions' own arguments instead. Both also take
> a large set of layout arguments, two-way faceting, row labels, legend
> geometry, reference lines; that are covered in
> `vignette("tutorial")` under *Publication layouts*.

## Structure plot

The **structure plot** shows posterior community membership probabilities for
each sample as stacked bars, the classic "STRUCTURE" visualization familiar
from population genetics.

```{r structure_basic}
# Basic structure plot (all samples, single panel)
eDNA_dmm_structure(fit)
```

```{r structure_faceted}
# Faceted by true community, sorted by depth within each panel
p_structure <- eDNA_dmm_structure(
  fit,
  metadata  = data$metadata,     # labelling columns, not the model covariates
  facet_var = "TrueCommunity",   # one panel per true community
  sort_var  = "Depth"            # sort samples by depth within each panel
)
p_structure
```

The structure plot is best read by looking for **vertical patterns** (samples
dominated by a single color) and **transitions** (where one community gives
way to another along the sorting axis). In the example dataset the four
communities are separated by depth and distance from shore, so sorting by
`Depth` within each true-community panel should give panels that are each
dominated by a single colour.

```{r structure_custom_colors}
# Use custom community colors
p_structure_custom <- eDNA_dmm_structure(
  fit,
  metadata         = data$metadata,
  facet_var        = "TrueCommunity",
  sort_var         = "Depth",
  community_colors = c(
    "Community 1" = "#E63946",
    "Community 2" = "#457B9D",
    "Community 3" = "#2A9D8F"
  )
)
p_structure_custom
```

```{r structure_save}
# Save at publication quality
ggplot2::ggsave(
  "structure_plot.png",
  p_structure,
  width  = 10,
  height = 4,
  dpi    = 300
)
```

```{r structure_customize}
# Further ggplot2 customization (e.g., change the strip text)
library(ggplot2)
p_structure +
  theme(
    strip.text = element_text(size = 14, face = "bold"),
    plot.title = element_text(size = 16)
  )
```
## Community compositions

Where the structure plot answers "which community does each *sample* belong
to?", `eDNA_dmm_compositions()` answers "what does each *community* look like
in species space?", the posterior mean composition of each community.

```{r compositions}
eDNA_dmm_compositions(
  fit,
  top_n           = 20,       # show the top 20 taxa; rest pooled into "Other"
  legend_position = "right"
)
```

Colours are shared with `plot_true_compositions()` and `eDNA_dmm_structure()`,
so the same taxon keeps the same colour across plots. That sharing is only
automatic when the plots are drawn from the same taxon set, if you put an
observed-composition plot next to a fitted one, rank the taxa once and pass the
same set to both:

```{r compositions_shared, eval=FALSE}
taxa <- dmm_taxon_order(fit, top_n = 20)

plot_true_compositions(data$counts, taxa_include = taxa)
eDNA_dmm_compositions(fit, top_n = 20)
```


## NMDS ordination

The **NMDS plot** shows community composition in ordination space. Points are
colored by their MAP community assignment, and size reflects assignment
certainty. Dashed ellipses encompass 95% of each community's points.

```{r nmds}
nmds_result <- eDNA_dmm_nmds(fit)

# Print the plot
nmds_result$plot

# Check NMDS stress (< 0.10 is good; > 0.20 may need k=3 dimensions)
message("NMDS stress: ", round(nmds_result$nmds$stress, 3))
```

Well-separated communities will appear as distinct point clusters. Overlapping
clusters indicate that the communities are compositionally similar and/or that
some samples are ambiguously assigned (which you will also see as smaller points,
since size reflects certainty).

```{r nmds_3d}
# Fit a 3D NMDS for lower stress, then view NMDS1 vs NMDS3
nmds_3d <- eDNA_dmm_nmds(fit, k = 3, nmds_axes = c(1, 3))
nmds_3d$plot
```

```{r nmds_custom}
# No ellipses, custom colors, Jaccard distance instead of Bray-Curtis
eDNA_dmm_nmds(
  fit,
  distance         = "jaccard",
  show_ellipse     = FALSE,
  community_colors = c("Community 1" = "#E63946",
                       "Community 2" = "#457B9D",
                       "Community 3" = "#2A9D8F")
)$plot
```

```{r nmds_save}
ggplot2::ggsave(
  "nmds.png",
  nmds_result$plot,
  width = 7,
  height = 6,
  dpi = 300
)
```

## Beta coefficient plots

The **beta plot** compares the posterior distribution of each softmax
regression coefficient to its prior. When the posterior is pulled away from
the prior, the data support that covariate's effect on community membership.

```{r beta_joint}
# Joint layout: communities overlaid per covariate
beta_result <- eDNA_dmm_beta(fit)
beta_result$plot
```

**Reading the beta plot:**

- **Grey** = the prior: Normal(0, 1). This is what the model assumed before
  seeing the data.
- **Red/Blue** = the posterior for each community (relative to Community 3,
  the reference).
- If the colored peak is shifted clearly away from grey (and from zero), that
  covariate has a genuine effect on community membership.
- The annotation shows the posterior mean, 90% CI, and P(direction), the
  probability that the coefficient has the sign it appears to have.

For the example data, you should see:
- **Latitude**: Community 1 posterior negative (prefers southern sites),
  Community 2 positive (prefers northern sites)
- **Depth**: Community 1 negative (prefers shallow), Community 2 positive
  (prefers deep)

```{r beta_separate}
# Separate layout: one row per community
eDNA_dmm_beta(fit, layout = "separate")$plot
```

```{r beta_table}
# Access the coefficient summary table
beta_result$table
# Columns: community, covariate, mean, ci_5, ci_95, prob_positive, prob_negative,
#          ess, reliability
```

The **reliability** column classifies each coefficient based on its
effective sample size (ESS):
- `"trustworthy"`: ESS > 400, use these with confidence
- `"cautious"`: ESS 100–400, treat as indicative, not definitive
- `"unreliable"`: ESS < 100, do not interpret; run longer chains

```{r beta_save}
ggplot2::ggsave(
  "beta_coefficients.png",
  beta_result$plot,
  width  = 8,
  height = 5,
  dpi    = 300
)
```

---

# Combining plots with cowplot and patchwork

The ggplot2 objects returned by all three visualization functions can be
combined freely. Here is an example using `cowplot`:

```{r cowplot, eval=FALSE}
install.packages("cowplot")
library(cowplot)

p1 <- eDNA_dmm_structure(fit, metadata = data$metadata,
                          facet_var = "TrueCommunity", sort_var = "Depth")
p2 <- eDNA_dmm_nmds(fit)$plot
p3 <- eDNA_dmm_beta(fit)$plot

combined <- cowplot::plot_grid(
  p2, p3,
  ncol   = 2,
  labels = c("A", "B"),
  label_size = 16
)
combined_full <- cowplot::plot_grid(
  p1, combined,
  ncol        = 1,
  rel_heights = c(1, 1.4),
  labels      = c("C", ""),
  label_size  = 16
)
cowplot::ggsave2("figure_1.png", combined_full, width = 14, height = 10, dpi = 300)
```

And with `patchwork` (another popular option):

```{r patchwork, eval=FALSE}
install.packages("patchwork")
library(patchwork)

(p2 | p3) / p1 +
  patchwork::plot_annotation(tag_levels = "A")
```

---

# Advanced topics

## Accessing the raw Stan fit

For diagnostics beyond what `summary()` provides, the raw
`rstan::stanfit` object is available in `fit$stan_fit`:

```{r stan_raw}
library(rstan)

# Full parameter summary
print(fit$stan_fit, pars = c("alpha", "pi[1,1]", "pi[2,1]", "beta[1,1]"))

# Traceplots
traceplot(fit$stan_fit, pars = "alpha")

# Pairs plot (joint posterior for a few parameters)
pairs(fit$stan_fit, pars = c("alpha", "beta[1,1]", "beta[1,2]"))
```

## Fixing alpha (known overdispersion)

If you have estimated the overdispersion parameter from a prior study or
pilot dataset, you can fix it by editing the Stan model to use the
"Option B" variant (see comments in `inst/stan/dmm.stan`). This reduces
the number of parameters and can improve convergence.

## Increasing iterations for better ESS

If `summary()` reports low ESS for some parameters:

```{r more_iter}
fit_long <- eDNA_dmm(
  counts     = data$counts,
  covariates = covariate_matrix,
  K          = 3,
  iter       = 10000,    # 10,000 total
  warmup     = 5000      # 5,000 warmup
)
```

## Silent fitting in loops

When fitting multiple K values manually, suppress output:

```{r silent}
results <- lapply(2:5, function(k) {
  eDNA_dmm(data$counts, covariate_matrix, K = k, verbose = FALSE)
})
names(results) <- paste0("K", 2:5)
```

---

# Troubleshooting

## "Divergent transitions" in diagnostics

Divergent transitions indicate that the sampler is having trouble exploring
the posterior geometry. Try:

1. Increase `adapt_delta` toward 0.99: `eDNA_dmm(..., adapt_delta = 0.99)`
2. Increase `max_treedepth` to 14 or 15
3. Reduce K, more communities require more complex geometry
4. Check that all-zero taxa have been removed from your count matrix

## Very low ESS for beta parameters

Low beta ESS often means the covariate has little effect and the posterior
is flat. This is not necessarily a problem; it means the data do not support
that covariate driving community structure. Check the beta plot: if the
posterior looks nearly identical to the prior, the data are uninformative
about that effect.

## "Only X stations pass min_station_reads" warning

This is not from `eDNA_dmm()` itself but from `validate_counts()`. Samples
with fewer than 100 total reads receive this warning. Remove low-coverage
samples: `counts <- counts[rowSums(counts) >= 1000, ]`

## Compilation takes a long time

Stan compiles the model to C++ code the first time. This takes 30–120 seconds
but only happens once. On subsequent calls, rstan uses the cached compiled
model (set `rstan_options(auto_write = TRUE)` to ensure caching).

## "object 'example_edna' not found"

Run `library(eDNAstructure)` first. The dataset is lazily loaded and becomes
available only after loading the package.

---

# Citation

If you use `eDNAstructure` in published research, please cite:


> Brandão-Dias et al. (year). Multinomial mixture models from environmental DNA reveal 
> depth stability and dynamic surface turnover of marine vertebrate communities. Under review
 
> Brandão-Dias et al. (year). eDNAstructure: Dirichlet-Multinomial Mixture Models for
> eDNA Metabarcoding Community Structure. R package version 0.1.0.
> https://github.com/pedrobdfp/eDNA_structure

Please also cite Stan:

> Carpenter B. et al. (2017). Stan: A probabilistic programming language.
> *Journal of Statistical Software*, 76(1).

---

# Session information

```{r session}
sessionInfo()
```
