---
title: "Getting Started with eDNAstructure"
subtitle: "Fitting Dirichlet-Multinomial Mixture Models to eDNA Metabarcoding Data"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 3
    number_sections: true
vignette: >
  %\VignetteIndexEntry{Getting Started with eDNAstructure}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse  = TRUE,
  comment   = "#>",
  eval      = FALSE,   # set to TRUE once package is installed with rstan
  fig.width = 7,
  fig.height = 5
)
```

# Introduction

`eDNAstructure` fits **Dirichlet-Multinomial Mixture (DMM) models** to
environmental DNA (eDNA) read count data from metabarcoding surveys. The
central question the model answers is:

> *Given read count data from many sampling locations, how many distinct
> ecological communities exist in my study area; and what environmental
> variables predict which community a sample belongs to?*

The model is appropriate whenever you have:

- A matrix of read counts (samples × taxa or samples × ASVs)
- Two or more sampling locations
- An interest in whether samples cluster into distinct assemblage types
- Optionally: environmental covariates you believe drive that clustering

This vignette walks you through the complete workflow from raw data to
publication-ready figures. All code uses the built-in example dataset so
you can run everything yourself.

---

# Installation

## Step 1: Install R

If you do not yet have R installed, download the latest version for your
operating system from <https://cran.r-project.org/>. Version 4.1.0 or later
is required.

We also strongly recommend installing **RStudio**, a free graphical interface
for R: <https://posit.co/download/rstudio-desktop/>.

## Step 2: Install Stan and rstan

`eDNAstructure` uses the Stan probabilistic programming language for Bayesian
inference. Stan has its own installation requirements beyond a standard R
package (and it is usually very annoying). Follow the official instructions carefully:

**<https://mc-stan.org/rstan/articles/rstan.html>**

The key steps are:

1. Install a C++ toolchain for your OS:
   - **Windows**: Install Rtools from <https://cran.r-project.org/bin/windows/Rtools/>
   - **macOS**: Install the Xcode Command Line Tools (`xcode-select --install` in Terminal)
   - **Linux**: Ensure `build-essential` (or equivalent) is installed

2. Install `rstan` from CRAN:
   ```r
   install.packages("rstan")
   ```

3. Verify the installation works:
   ```r
   library(rstan)
   example(stan_model, package = "rstan", run.dontrun = TRUE)
   ```
   If this runs without errors and produces output like `"SAMPLING FOR MODEL..."`,
   Stan is working correctly.

## Step 3: Install eDNAstructure

Install directly from GitHub using the `remotes` package:

```r
# Install remotes if you don't have it
install.packages("remotes")

# Install eDNAstructure
remotes::install_github("pedrobdfp/eDNA_structure")
```

This will also install the required R package dependencies automatically:
`ggplot2`, `dplyr`, `tidyr`, `vegan`, `posterior`, `loo`, and `scales`.

## Step 4: Load the package

```{r load}
library(eDNAstructure)
```

---

# Your data and what the model expects

Before fitting a model, it helps to understand exactly what format the
functions expect. We will use the built-in example dataset to illustrate.
The model needs **two things**, and only the first is required:

1. **A species/ASV × site table**: the count matrix.
2. **A covariate table**: *optional*. Supply it only if you want to model how
   environmental variables drive community membership. `eDNA_dmm()` runs
   perfectly well without one.

## Starting from one file

In practice your data usually arrive as a single spreadsheet with sample IDs,
environmental variables and counts all side by side. A file of exactly that
shape ships with the package, so the split is easy to see:

```{r raw_file}
path <- system.file("extdata", "example_edna_raw.csv", package = "eDNAstructure")
raw  <- read.csv(path, check.names = FALSE)

raw[1:3, 1:7]
```

Split it into the two tables the model wants:

```{r raw_split}
# Which columns hold taxa? Adjust this one line to match your own file --
# a name prefix, a known column range, or setdiff() against your metadata names.
taxon_cols <- grep("^Sp_", names(raw))

# 1. THE SPECIES/ASV x SITE TABLE -- required
counts <- as.matrix(raw[, taxon_cols])
rownames(counts) <- raw$sample_id
storage.mode(counts) <- "integer"

# 2. THE COVARIATE TABLE -- optional, numeric columns only
covariates <- raw[, c("Depth", "Distance_shore")]
rownames(covariates) <- raw$sample_id

dim(counts)
dim(covariates)
```

Both fits are valid:

```{r raw_split_fit, eval=FALSE}
fit <- eDNA_dmm(counts, K = 4)                           # without covariates
fit <- eDNA_dmm(counts, covariates = covariates, K = 4)  # with them
```

> **The covariate table must be numeric, and only numeric.** `eDNA_dmm()` fits
> every column you hand it, so an identifier such as `sample_id`, or a label
> such as `TrueCommunity`, must not be in it, the model would try to fit the
> ID as an environmental gradient. Keep those in a separate metadata frame and
> pass it to the plotting functions' `metadata` argument instead. This is why
> `get_example_data()` returns `covariates` and `metadata` as two objects.

## The count matrix

The most important input is a **sample × taxon count matrix**. The function
`eDNA_dmm()` accepts:

- A `matrix` with non-negative integer values
- Or a `data.frame` where every column is numeric

Rows represent **samples** (individual water samples or stations); columns
represent **taxa or ASVs**, taxonomic annotation is not required. The model is
equally applicable to raw ASV tables and to taxonomically-collapsed count
tables.

```{r count_matrix}
# Load the example dataset
data <- get_example_data()

# Inspect the count matrix
dim(data$counts)  # 20 samples x 32 taxa

# First 5 samples, first 6 taxa
data$counts[1:5, 1:6]
```

The output will look something like:
```
        Sp_1 Sp_2 Sp_3 Sp_4 Sp_7 Sp_9
STN_001 2054 1794  780  717 1087  454
STN_002  620 3256 1492    0    0 1359
STN_003  192 3306    0    0    0  906
STN_004  621  877  401 1305    0  590
STN_005 4066   57    0    0  492    0
```

Each number is the count of sequencing reads assigned to that taxon in that
sample. These are raw counts, **do not** normalize to proportions before
fitting. The model handles composition estimation internally.

**Key rules for the count matrix:**

| Requirement | What happens if violated |
|-------------|--------------------------|
| All values ≥ 0 | Error with informative message |
| No NA values | Error with informative message |
| No all-zero rows | Error: empty samples cannot be modelled |
| At least 2 columns | Error |
| Integer values | Warning, then automatic coercion |
| Row sums ≥ 100 | Warning (very low counts = unreliable compositions) |

## The covariate data frame

The second input (optional) is a **sample × covariate** matrix or data frame.
If you pass covariates, the model will estimate how each covariate shifts the
probability that a sample belongs to each community.

```{r covariates}
# Inspect the covariate data frame
head(data$covariates)   # numeric only
```

```
        Depth Distance_shore
STN_001  83.6          137.0
STN_002  63.7          137.0
STN_003  66.8          148.0
```

`covariates` holds exactly these two numeric columns, nothing else. There is
no `sample_id` or `TrueCommunity` column to drop; that separation is done for
you already (see the labelling-column note above). If you ever see extra
columns here, or an "object 'example_edna' not found"-style mismatch after
loading a saved workspace, restart R without restoring old objects, an
object left over from an earlier session can shadow the package's own data.

**Critical:** the covariate data frame must have exactly the same number of
rows as your count matrix, **in the same order**. Each row `i` of the
covariate data frame corresponds to row `i` of the count matrix. If these
are misaligned, the model will silently use the wrong covariate values for
each sample.

A safe way to ensure alignment is to match on sample IDs:

```{r align}
# Ensure both objects are sorted by sample ID before fitting
counts_sorted      <- data$counts[order(rownames(data$counts)), ]
covariates_sorted  <- data$covariates[order(rownames(data$covariates)), ]
metadata_sorted    <- data$metadata[order(data$metadata$sample_id), ]

# Verify alignment: counts and covariates share row names; metadata carries
# its own sample_id column
stopifnot(all(rownames(counts_sorted) == rownames(covariates_sorted)))
stopifnot(all(rownames(counts_sorted) == metadata_sorted$sample_id))
```

### Which covariates to include?

Include any continuous environmental variable you hypothesize might predict
community membership:

- **Spatial**: latitude, longitude, distance from shore, distance from canyon
- **Physical**: depth, temperature, salinity, oxygen, turbidity
- **Derived**: distance from thermocline, distance to seafloor, upwelling index

Covariates are Z-score standardized internally by default (`scale_covariates = TRUE`).
This means the fitted beta coefficients are directly comparable across
covariates regardless of their original units. Do not standardize your
covariates yourself unless you set `scale_covariates = FALSE`.

Categorical variables (e.g., site, year) can be included as numeric dummy
indicators, but interpret with care: the model treats them as continuous.
For categorical grouping it is often better to use them for faceting in
plots (post-hoc) rather than as covariates.

### Preparing a long-format table

If your data come in long format (one row per sample-taxon combination),
convert to wide format before using this package:

```{r long_to_wide, eval=FALSE}
library(tidyr)
library(dplyr)

# Assume `long_df` has columns: sample_id, taxon, reads
count_matrix <- long_df %>%
  pivot_wider(
    names_from  = taxon,
    values_from = reads,
    values_fill = 0          # replace missing with 0
  ) %>%
  tibble::column_to_rownames("sample_id") %>%
  as.matrix()
```

---

# Fitting the model

## Basic fit: K = 2

Start simple. A K = 2 model answers: "can my samples be divided into two
distinct communities?"

```{r fit_k2}
# Select which covariates to use
covariate_matrix <- data$covariates   # already numeric-only

# Fit a K=2 DMM
fit_k2 <- eDNA_dmm(
  counts     = data$counts,
  covariates = covariate_matrix,
  K          = 2
)
```

When you run this, you will see Stan's sampling progress printed to the
console. This is normal. Four chains of 4000 iterations (2000 warmup)
typically takes 1–5 minutes depending on N and S.

```r
## Fitting DMM: N=20 samples, S=32 taxa, K=2 communities, P=2 covariates
## MCMC: 4 chains on 4 cores, 4000 iterations (2000 warmup)
## ...
## Chain 1: Iteration: 1 / 4000 [  0%]  (Warmup)
## Chain 1: Iteration: 400 / 4000 [ 10%]  (Warmup)
## ...
## Chain 4: Elapsed Time: ...
```

### Why four chains? What about label switching?

Mixture models have a fundamental ambiguity: "Community 1" in one chain might
correspond to "Community 2" in another (label switching), which would make
naive between-chain Rhat look terrible even when every chain has converged
individually. `eDNA_dmm()` handles this for you: after sampling, it aligns
every draw from every chain onto a common labelling (via the `label.switching`
package's STEPHENS algorithm) before computing any diagnostic. You do not need
to work around label switching yourself, run four chains as usual, `print()`
and `summary()` report Rhat/ESS computed *after* alignment.

`fit$alignment$diagnostics` (or the "Chain agreement" line in `summary()`)
additionally tells you whether the chains agree on the *same* solution, not
merely the same names for it. Low agreement across chains means the chains
found genuinely different partitions of the data, more sampling will not fix
that; it means K is likely misspecified for the amount of signal available
(see the box below).

## Examining the results

```{r print_fit}
# Quick overview
print(fit_k2)
```

```
eDNA Dirichlet-Multinomial Mixture Model
=========================================
  K (communities)  : 2
  N (samples)      : 20
  S (taxa)         : 32
  Covariates       : Depth, Distance_shore
  Mean alpha       : 1.94
  Chains           : 4, labels aligned by STEPHENS
  Chain agreement  : 0.500 (lp Rhat 1.529)

Community sizes (most probable assignment):
  Community 1     : 10 samples (50.0%)
  Community 2     : 10 samples (50.0%)
```

```{r summary_fit}
# Full diagnostics
summary(fit_k2)
```

> **This particular fit is a good example of a warning sign, not a clean
> result.** The example dataset actually has four true communities. Forcing
> K = 2 makes the model merge pairs of them, and the diagnostics say so
> plainly: chain agreement is only 50%, and both `pi` and `beta` have poor
> Rhat/ESS after alignment. That is not a bug in the model or the data, it is
> the model correctly reporting that two clusters do not describe this survey
> well. Rather than hide it, use it: this is exactly the kind of signal that
> motivates the LOO-based K-selection workflow later in this vignette. A
> well-specified K (see the K = 4 fit below, once LOO has selected it)
> converges cleanly, with chain agreement near 100% and Rhat close to 1.

The summary prints:

1. **Alpha (overdispersion)**: typical values for eDNA data are 1–5.
   High alpha (>10) means counts are nearly multinomial (low overdispersion).
   Low alpha (~1) means very high overdispersion, typical for eDNA.

2. **Label alignment and chain agreement**: whether the four chains describe
   the same partition of the samples (not just whether their labels match).
   Near 100% means one solution was found repeatedly; well below that means
   the chains disagree about the grouping itself.

3. **Assignment certainty**: the fraction of samples with ≥80% posterior
   probability of belonging to one community. High certainty (>90%) means the
   communities are well-separated.

4. **Covariate effects (beta)**: each community's coefficient is its
   *deviation from the average community* for that covariate, there is no
   reference community (see "Beta coefficient plots" below for why). A 90% CI
   that excludes 0 is evidence for a real effect; check the `reliability`
   column (ESS-based) before trusting any single coefficient.

5. **HMC diagnostics**: look for "No divergences", "No max treedepth",
   "No low BFMI". Any warnings here should be addressed before interpreting
   results.

## Accessing model components directly

The fit object is a standard R list. You can access all components directly:

```{r access}
# Posterior community membership probabilities (one row per sample)
head(fit_k2$sample_info)
# Columns: sample_id, prob_comm1, prob_comm2, z_hat, assignment_certainty

# Posterior mean community compositions (K x S matrix)
fit_k2$pi_mean          # rows = communities, columns = taxa
round(fit_k2$pi_mean[, 1:5], 3)

# Beta coefficient posterior summaries
fit_k2$beta_summary

# Posterior mean alpha
fit_k2$alpha_mean
```


# Choosing K: How many communities?

In practice, you rarely know K in advance. The recommended approach is to fit
models for K = 2 through K = 6 (or however many is ecologically plausible)
and compare them using Leave-One-Out cross-validation (LOO-CV).

## Using eDNA_loo()

```{r loo}
# Fit K=2 through K=5 and compare (this takes several minutes)
loo_result <- eDNA_loo(
  counts     = data$counts,
  covariates = covariate_matrix,
  K_range    = 2:5,
  iter       = 4000,
  warmup     = 2000
)
```

```{r loo_plot}
# Elbow plot
loo_result$plot
```

```{r loo_table}
# LOO-ELPD table
loo_result$loo_table
```

```
  K       elpd        se
1 2 -1204.3  81.3
2 3 -1181.6  78.9
3 4 -1148.2  74.1   <- gain from K=3 to K=4 is large
4 5 -1146.9  75.8   <- K=4 to K=5 is within ~2 SE: negligible gain
```

**Reading the elbow plot:**

- **Higher LOO-ELPD** = better predictive performance = better model
- The "elbow" is the K beyond which improvements are negligible
- If two models are within ~2 SE of each other, prefer the simpler one (lower K)
- Biologically, K should also be interpretable, resist the urge to over-split

For the example data (which was generated with K = 4), the LOO-ELPD should
peak around K = 4 before levelling off. The model successfully recovers the
true number of communities.

> **Want more than one plot?** `eDNA_dmm_k_diagnostics(loo_result$fits, K_values = 2:5, elpd_by_run
> = loo_result$loo_by_chain, convergence = loo_result$loo_table)` draws six panels side by side,
> predictive fit, reliability, chain convergence, assignment certainty and community distinctness,
> so a K that looks good on the elbow plot alone but is poorly identified or uninterpretable is easy
> to catch. See `vignette("tutorial")` for the full walkthrough, or `?eDNA_dmm_k_diagnostics`.

## Fitting the selected K

Once you have chosen K, fit the final model with more iterations for better
posterior summaries:

```{r fit_final}
fit <- eDNA_dmm(
  counts     = data$counts,
  covariates = covariate_matrix,
  K          = 4,
  iter       = 4000,
  warmup     = 2000
)
```

---

# Visualization

All four visualization functions return `ggplot2` objects. This means you
can:

1. Print them directly to see the plot
2. Save them with `ggsave()`
3. Modify them by adding ggplot2 layers
4. Combine them with `cowplot::plot_grid()` or `patchwork`

> **One exception.** `eDNA_dmm_structure()` and `plot_true_compositions()`
> return a *cowplot grid* object rather than a ggplot when `facet_row_var` is
> supplied, because they build one sub-plot per row level to keep panel widths
> proportional to sample count. Adding ggplot2 layers with `+` will not work on
> those; style them through the functions' own arguments instead. Both also take
> a large set of layout arguments, two-way faceting, row labels, legend
> geometry, reference lines; that are covered in
> `vignette("tutorial")` under *Publication layouts*.

## Structure plot

The **structure plot** shows posterior community membership probabilities for
each sample as stacked bars, the classic "STRUCTURE" visualization familiar
from population genetics.

```{r structure_basic}
# Basic structure plot (all samples, single panel)
eDNA_dmm_structure(fit)
```

```{r structure_faceted}
# Faceted by true community, sorted by depth within each panel
p_structure <- eDNA_dmm_structure(
  fit,
  metadata  = data$metadata,     # labelling columns, not the model covariates
  facet_var = "TrueCommunity",   # one panel per true community
  sort_var  = "Depth"            # sort samples by depth within each panel
)
p_structure
```

The structure plot is best read by looking for **vertical patterns** (samples
dominated by a single color) and **transitions** (where one community gives
way to another along the sorting axis). In the example dataset the four
communities are separated by depth and distance from shore, so sorting by
`Depth` within each true-community panel should give panels that are each
dominated by a single colour.

```{r structure_custom_colors}
# Use custom community colors
p_structure_custom <- eDNA_dmm_structure(
  fit,
  metadata         = data$metadata,
  facet_var        = "TrueCommunity",
  sort_var         = "Depth",
  community_colors = c(
    "Community 1" = "#E63946",
    "Community 2" = "#457B9D",
    "Community 3" = "#2A9D8F",
    "Community 4" = "#E9C46A"
  )
)
p_structure_custom
```

```{r structure_save}
# Save at publication quality
ggplot2::ggsave(
  "structure_plot.png",
  p_structure,
  width  = 10,
  height = 4,
  dpi    = 300
)
```

```{r structure_customize}
# Further ggplot2 customization (e.g., change the strip text)
library(ggplot2)
p_structure +
  theme(
    strip.text = element_text(size = 14, face = "bold"),
    plot.title = element_text(size = 16)
  )
```
## Community compositions

Where the structure plot answers "which community does each *sample* belong
to?", `eDNA_dmm_compositions()` answers "what does each *community* look like
in species space?", the posterior mean composition of each community.

```{r compositions}
eDNA_dmm_compositions(
  fit,
  top_n           = 20,       # show the top 20 taxa; rest pooled into "Other"
  legend_position = "right"
)
```

Colours are shared with `plot_true_compositions()` and `eDNA_dmm_structure()`,
so the same taxon keeps the same colour across plots. That sharing is only
automatic when the plots are drawn from the same taxon set, if you put an
observed-composition plot next to a fitted one, rank the taxa once and pass the
same set to both:

```{r compositions_shared, eval=FALSE}
taxa <- dmm_taxon_order(fit, top_n = 20)

plot_true_compositions(data$counts, taxa_include = taxa)
eDNA_dmm_compositions(fit, top_n = 20)
```


## NMDS ordination

The **NMDS plot** shows community composition in ordination space. Points are
colored by their MAP community assignment, and size reflects assignment
certainty. Dashed ellipses encompass 95% of each community's points.

```{r nmds}
nmds_result <- eDNA_dmm_nmds(fit)

# Print the plot
nmds_result$plot

# Check NMDS stress (< 0.10 is good; > 0.20 may need k=3 dimensions)
message("NMDS stress: ", round(nmds_result$nmds$stress, 3))
```

Well-separated communities will appear as distinct point clusters. Overlapping
clusters indicate that the communities are compositionally similar and/or that
some samples are ambiguously assigned (which you will also see as smaller points,
since size reflects certainty).

```{r nmds_3d}
# Fit a 3D NMDS for lower stress, then view NMDS1 vs NMDS3
nmds_3d <- eDNA_dmm_nmds(fit, k = 3, nmds_axes = c(1, 3))
nmds_3d$plot
```

```{r nmds_custom}
# No ellipses, custom colors, Jaccard distance instead of Bray-Curtis
eDNA_dmm_nmds(
  fit,
  distance         = "jaccard",
  show_ellipse     = FALSE,
  community_colors = c("Community 1" = "#E63946",
                       "Community 2" = "#457B9D",
                       "Community 3" = "#2A9D8F",
                       "Community 4" = "#E9C46A")
)$plot
```

```{r nmds_save}
ggplot2::ggsave(
  "nmds.png",
  nmds_result$plot,
  width = 7,
  height = 6,
  dpi = 300
)
```

## Beta coefficient plots

The **beta plot** compares the posterior distribution of each softmax
regression coefficient to its prior. When the posterior is pulled away from
the prior, the data support that covariate's effect on community membership.

```{r beta_joint}
# Joint layout: communities overlaid per covariate
beta_result <- eDNA_dmm_beta(fit)
beta_result$plot
```

**Reading the beta plot:**

- **Grey** = the prior: Normal(0, 1). This is what the model assumed before
  seeing the data.
- **Coloured** = the posterior for each community. `eDNA_dmm()` uses a
  **centred** softmax parameterization, there is no reference community.
  Each community's coefficient for a covariate is its *deviation from the
  average community* on that covariate (the coefficients across communities
  sum to zero for each covariate). To read one community as a contrast
  against another, subtract that community's posterior draws from the rest.
- If the coloured peak is shifted clearly away from grey (and from zero), that
  covariate has a genuine effect on community membership.
- The annotation shows the posterior mean, 90% CI, and P(direction), the
  probability that the coefficient has the sign it appears to have.

For the example data (four communities defined by Depth x Distance_shore,
means 70/40 m and 140/100 units), you should see:
- **Depth**: Communities 1 and 3 (deep, mean 70) positive, Communities 2 and 4
  (shallow, mean 40) negative
- **Distance_shore**: Communities 1 and 2 (offshore, mean 140) positive,
  Communities 3 and 4 (inshore, mean 100) negative

```{r beta_separate}
# Separate layout: one row per community
eDNA_dmm_beta(fit, layout = "separate")$plot
```

```{r beta_table}
# Access the coefficient summary table
beta_result$table
# Columns: community, covariate, mean, ci_5, ci_95, prob_positive, prob_negative,
#          ess, reliability
```

The **reliability** column classifies each coefficient based on its
effective sample size (ESS):
- `"trustworthy"`: ESS > 400, use these with confidence
- `"cautious"`: ESS 100–400, treat as indicative, not definitive
- `"unreliable"`: ESS < 100, do not interpret; run longer chains

```{r beta_save}
ggplot2::ggsave(
  "beta_coefficients.png",
  beta_result$plot,
  width  = 8,
  height = 5,
  dpi    = 300
)
```

---

# Combining plots with cowplot and patchwork

The ggplot2 objects returned by all three visualization functions can be
combined freely. Here is an example using `cowplot`:

```{r cowplot, eval=FALSE}
install.packages("cowplot")
library(cowplot)

p1 <- eDNA_dmm_structure(fit, metadata = data$metadata,
                          facet_var = "TrueCommunity", sort_var = "Depth")
p2 <- eDNA_dmm_nmds(fit)$plot
p3 <- eDNA_dmm_beta(fit)$plot

combined <- cowplot::plot_grid(
  p2, p3,
  ncol   = 2,
  labels = c("A", "B"),
  label_size = 16
)
combined_full <- cowplot::plot_grid(
  p1, combined,
  ncol        = 1,
  rel_heights = c(1, 1.4),
  labels      = c("C", ""),
  label_size  = 16
)
cowplot::ggsave2("figure_1.png", combined_full, width = 14, height = 10, dpi = 300)
```

And with `patchwork` (another popular option):

```{r patchwork, eval=FALSE}
install.packages("patchwork")
library(patchwork)

(p2 | p3) / p1 +
  patchwork::plot_annotation(tag_levels = "A")
```

---

# Advanced topics

## Accessing the raw Stan fit

For diagnostics beyond what `summary()` provides, the raw
`rstan::stanfit` object is available in `fit$stan_fit`:

```{r stan_raw}
library(rstan)

# Full parameter summary
print(fit$stan_fit, pars = c("alpha", "pi[1,1]", "pi[2,1]", "beta[1,1]"))

# Traceplots
traceplot(fit$stan_fit, pars = "alpha")

# Pairs plot (joint posterior for a few parameters)
pairs(fit$stan_fit, pars = c("alpha", "beta[1,1]", "beta[1,2]"))
```

## Fixing alpha (known overdispersion)

If you have estimated the overdispersion parameter from a prior study or
pilot dataset, you can fix it by editing `inst/stan/dmm.stan` directly:
replace the `alpha ~ gamma(alpha_shape, alpha_rate)` prior and the
`real<lower=0> alpha;` parameter declaration with a fixed `real` value in the
`data` block instead. This removes one parameter from the model and can
improve convergence when the true value is well known. There is no bundled
"fixed-alpha" variant, this requires editing and recompiling the model
yourself (`eDNA_clear_stan_cache()` after editing, so the change is picked up).

## Increasing iterations for better ESS

If `summary()` reports low ESS for some parameters:

```{r more_iter}
fit_long <- eDNA_dmm(
  counts     = data$counts,
  covariates = covariate_matrix,
  K          = 4,
  iter       = 10000,    # 10,000 total
  warmup     = 5000      # 5,000 warmup
)
```

## Silent fitting in loops

When fitting multiple K values manually, suppress output:

```{r silent}
results <- lapply(2:5, function(k) {
  eDNA_dmm(data$counts, covariate_matrix, K = k, verbose = FALSE)
})
names(results) <- paste0("K", 2:5)
```

---

# Troubleshooting

## "Divergent transitions" in diagnostics

Divergent transitions indicate that the sampler is having trouble exploring
the posterior geometry. Try:

1. Increase `adapt_delta` toward 0.99: `eDNA_dmm(..., adapt_delta = 0.99)`
2. Increase `max_treedepth` to 14 or 15
3. Reduce K, more communities require more complex geometry
4. Check that all-zero taxa have been removed from your count matrix

## Very low ESS for beta parameters

Low beta ESS often means the covariate has little effect and the posterior
is flat. This is not necessarily a problem; it means the data do not support
that covariate driving community structure. Check the beta plot: if the
posterior looks nearly identical to the prior, the data are uninformative
about that effect.

## "Only X stations pass min_station_reads" warning

This is not from `eDNA_dmm()` itself but from `validate_counts()`. Samples
with fewer than 100 total reads receive this warning. Remove low-coverage
samples: `counts <- counts[rowSums(counts) >= 1000, ]`

## Compilation takes a long time

Stan compiles the model to C++ code the first time. This takes 30–120 seconds
but only happens once. On subsequent calls, rstan uses the cached compiled
model (set `rstan_options(auto_write = TRUE)` to ensure caching).

## "object 'example_edna' not found"

Run `library(eDNAstructure)` first. The dataset is lazily loaded and becomes
available only after loading the package.

---

# Citation

If you use `eDNAstructure` in published research, please cite:


> Brandão-Dias et al. (year). Multinomial mixture models from environmental DNA reveal 
> depth stability and dynamic surface turnover of marine vertebrate communities. Under review
 
> Brandão-Dias et al. (year). eDNAstructure: Dirichlet-Multinomial Mixture Models for
> eDNA Metabarcoding Community Structure. R package version 0.1.0.
> https://github.com/pedrobdfp/eDNA_structure

Please also cite Stan:

> Carpenter B. et al. (2017). Stan: A probabilistic programming language.
> *Journal of Statistical Software*, 76(1).

---

# Session information

```{r session}
sessionInfo()
```
