---
title: "eDNAstructure: Full Tutorial"
subtitle: "From Simulation to Fitted Communities, A Complete Walkthrough"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 4
    number_sections: true
vignette: >
  %\VignetteIndexEntry{eDNA_structure: Full Tutorial}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse   = TRUE,
  comment    = "#>",
  eval       = FALSE,
  fig.width  = 8,
  fig.height = 5,
  message    = FALSE,
  warning    = FALSE
)
```

```{r load}
library(eDNAstructure)
library(ggplot2)    # for post-hoc plot customization
library(dplyr)      # for data manipulation
library(cowplot)    # for multi-panel figure assembly
```

---

# Overview

This tutorial walks through the complete `eDNAstructure` workflow from start to
finish. By the end you will know how to:

1. **Simulate** a realistic eDNA survey with known ground truth
2. **Inspect** raw community signal before fitting
3. **Format** data correctly for `eDNA_dmm()`
4. **Select K** using LOO cross-validation
5. **Fit** the Dirichlet-Multinomial Mixture model
6. **Visualize** community assignments, compositions, NMDS ordination, and
   covariate effects
7. **Verify** recovery of true parameters

The package implements a Bayesian Dirichlet-Multinomial Mixture (DMM) model.
Each sample is modeled as arising from a mixture of K latent communities, where
each community has a characteristic species composition (the π vectors) and
environmental covariates predict community membership through a softmax
regression (the β coefficients). The global overdispersion parameter α absorbs
both ecological and technical variance. Marginalizing over the discrete latent
community assignment makes inference exact; there is no variational
approximation.

For a shorter introduction, see `vignette("eDNA_structure")`.

---

# The simulation pipeline

`eDNAstructure` includes a full mechanistic simulation pipeline so you can
generate datasets with known community structure and verify that the model
recovers it. The same pipeline was used to generate the built-in example
dataset returned by `get_example_data()`.

## Why simulate?

Working with simulated data first has several practical advantages:

- **Controlled validation**: you know the true K, true compositions, and true
  covariate effects, so you can check whether the model gives the right answers
- **Power analysis**: you can vary the number of samples, amount of
  compositional overlap (spillover), or sequencing depth to understand what
  conditions the model requires to work well
- **Teaching**: students can see the ground truth alongside the fitted model,
  making it easy to build intuition about what "good recovery" looks like

## The five simulation steps

The pipeline runs five sequential steps. You can call them individually for
full control, or use `simulate_eDNA_survey()` to run all five at once.

### Step 1: Community compositions

`generate_community_compositions()` builds a K x S matrix where each row is
one community's species frequency vector. The key design choices are how many
dominant species each community has, whether communities share habitat groups
(which determines which species they share), and how much spillover there is
between communities.

```{r sim_compositions}
# Four communities, 40 species
# Communities 1+3 share a "deep habitat" group; 2+4 share a "surface habitat" group
community_mat <- generate_community_compositions(
  n_communities         = 4,
  n_species             = 40,
  n_dominant_range      = c(2, 3),        # 2-3 dominant species per community
  n_unique_low_freq     = 4,              # 4 species unique to each community
  community_groups      = c(1, 2, 1, 2), # depth groups: 1=deep, 2=surface
  n_group_shared        = 4,             # species shared within each depth group
  group_shared_presence = 1.0,
  dominant_freq         = c(0.08, 0.14),
  low_unique_freq       = c(0.02, 0.04),
  group_shared_freq     = c(0.02, 0.04),
  shared_freq           = c(0.02, 0.04),
  shared_presence       = 10,            # no globally ubiquitous species
  spillover             = 0.15,          # 15% of each species' frequency leaks elsewhere
  seed                  = 2026
)

dim(community_mat)            # 4 x 40
round(rowSums(community_mat), 6)   # each row sums to 1
```

The `spillover` parameter controls how exclusive each community's species pool
is. `spillover = 0` gives perfectly diagnostic communities where each species
is present only in one community. `spillover = 0.15` is more realistic: species
occur in multiple communities but at different relative abundances. Values above
~0.4 typically make communities hard to separate.

### Step 2: Contributors (organisms)

`generate_contributors()` assigns individual organisms to each sample by
drawing species identities from the community frequency vector. Body size and
distance from sampler are also drawn here because they affect eDNA shedding
in the next step.

```{r sim_contributors}
contrib_obj <- generate_contributors(
  community_compositions = community_mat,
  samples_per_community  = 5,           # 5 stations per community = 20 samples total
  n_contributors_range   = c(20, 100),  # 20-100 organisms contribute eDNA per sample
  n_distribution         = "Negative Binomial",  # overdispersed, more realistic than Poisson
  size_range             = c(1, 10),    # body size range (affects eDNA shedding rate)
  distance_range         = c(0, 100),   # distance from sampler (affects eDNA decay)
  seed                   = 2026
)

length(contrib_obj$contributors_list)   # 20 samples
head(contrib_obj$contributors_list[[1]])
```

### Step 3: eDNA shedding and sub-sampling

`generate_eDNA()` models how eDNA molecules move from organisms into the water
column and how many end up captured in a sampling bottle. Shedding scales
allometrically with body size (larger organisms shed more eDNA), decays
exponentially with distance from the sampler, and has lognormal noise that
represents individual variation in shedding rates.

```{r sim_edna}
eDNA_obj <- generate_eDNA(
  contributors_list = contrib_obj$contributors_list,
  shedding_rate     = 1000,    # baseline molecules shed per unit body size
  beta              = 0.75,    # allometric exponent (metabolic scaling)
  decay_rate        = 0.1,     # exponential distance decay constant
  shedding_error    = 0.3,     # lognormal noise SD on individual shedding
  bottle_volume     = 0.1,     # fraction of local eDNA pool captured per bottle
  bio_reps          = 1        # one bottle per station
)

head(eDNA_obj$Summary)
```

### Step 4: Metabarcoding

`simulate_metabarcoding()` applies amplification stochasticity and sequencing
noise to produce final integer read counts.

```{r sim_metab}
metab_df <- simulate_metabarcoding(
  eDNA_obj,
  mean_read_depth = 10000,   # mean reads per sample (typical for 16S amplicon)
  read_depth_sd   = 0.2,     # lognormal SD for between-sample read depth variation
  error_sd        = 0.05,    # Gaussian noise on species frequencies (amplification bias)
  rep             = 1        # one sequencing replicate per bottle
)

str(metab_df)
```

### Step 5: Environmental covariates

`generate_sample_covariates()` draws realistic covariate values for each
sample from community-specific distributions. Each community occupies a
distinct region of covariate space.

```{r sim_covariates}
# Define where each community lives in environmental space. The gap between
# community means is chosen to be a moderate multiple of the SD (~5): enough
# for a clean visual separation, but not so extreme that the covariates
# become (quasi-)perfectly separating in the softmax regression below, which
# would starve beta of gradient and produce poor chain mixing. Community
# means also stay well clear of 0, so sampling noise cannot produce a
# physically impossible negative Depth.
community_covariate_means <- matrix(
  c(70, 140,   # Community 1: deep + offshore
    40, 140,   # Community 2: shallow + offshore
    70, 100,   # Community 3: deep + inshore
    40, 100),  # Community 4: shallow + inshore
  nrow = 4, byrow = TRUE,
  dimnames = list(NULL, c("Depth", "Distance_shore"))
)

covariate_sds <- matrix(
  c(6, 8),               # Depth SD = 6, Distance_shore SD = 8, all communities
  nrow = 4, ncol = 2, byrow = TRUE,
  dimnames = list(NULL, c("Depth", "Distance_shore"))
)

sample_metadata <- generate_sample_covariates(
  contributors_list    = contrib_obj$contributors_list,
  community_covariates = community_covariate_means,
  covariate_sds        = covariate_sds
)

head(sample_metadata)
```

## Running the full pipeline at once

For most workflows, `simulate_eDNA_survey()` runs all five steps in one call
and returns a count matrix and covariate data frame ready for `eDNA_dmm()`.

```{r sim_full}
sim <- simulate_eDNA_survey(
  n_communities             = 4,
  n_species                 = 40,
  samples_per_community     = 5,
  community_covariate_means = community_covariate_means,
  covariate_sds             = covariate_sds,
  mean_read_depth           = 10000,
  spillover                 = 0.15,
  shedding_error            = 0.3,
  seed                      = 42
)

dim(sim$counts)       # 20 x S; some of the 40 species end up with zero reads
                       # across every sample and are dropped, so S < 40
head(sim$covariates)  # Depth, Distance_shore -- numeric only, straight into eDNA_dmm()
head(sim$metadata)    # sample_id, TrueCommunity, Depth, Distance_shore -- for plotting
```

---

# Data format

Before fitting, make sure your data are in the right format.

## The count matrix

The count matrix is the primary input to `eDNA_dmm()`.

```{r data_format_counts}
# Rows = samples, columns = taxa, values = integer read counts
sim$counts[1:4, 1:6]

# Check read depth distribution: large variation in read depth is normal
range(rowSums(sim$counts))
hist(rowSums(sim$counts), main = "Read depth distribution", xlab = "Total reads")

# Zero-count columns are dropped automatically with a warning, but best to
# handle this yourself so you know what is being removed
counts <- sim$counts[, colSums(sim$counts) > 0]
dim(counts)
```

**Key rules:**

- All values must be non-negative integers (or coercible to integers)
- No `NA` values, replace with 0 if a taxon was absent
- No all-zero rows, remove samples with zero total reads
- Row names are used as sample IDs throughout all plots; set them meaningfully
- Do **not** normalize or rarefy, the model handles read depth internally

## The covariate data frame

```{r data_format_covs}
# The covariate data frame must be in the same row order as the count matrix
head(sim$covariates)

# Always verify alignment before fitting
stopifnot(all(rownames(counts) == sim$metadata$sample_id))

# Select only the covariates you want to include in the model
cov_matrix <- sim$covariates   # already numeric-only
head(cov_matrix)
```

Covariates are Z-score standardized internally when `scale_covariates = TRUE`
(the default). Do not pre-standardize, the function handles this so all β
coefficients remain on a common scale and are directly comparable.

---

# Visualizing raw compositions (before fitting)

Before running the model, always visualize the raw species composition of each
sample. `plot_true_compositions()` produces a stacked bar plot in the same
layout as the structure and community composition plots, making side-by-side
comparison straightforward.

This step serves two purposes: it gives you a visual check that there is some
community structure in the data worth modeling, and it produces panel A of the
publication figure.

```{r plot_raw_unfaceted}
# All samples, no grouping: useful as a first look at any dataset
plot_true_compositions(counts, legend_nrow = 8)
```

```{r plot_raw_faceted}
# Faceted by true community: possible with simulated data where true labels
# are known. With real data, facet by a hypothesized grouping (site, season,
# depth stratum) to evaluate whether it explains the composition signal.
p_raw <- plot_true_compositions(
  counts,
  metadata        = sim$metadata,
  facet_var       = "TrueCommunity",
  sort_var        = "Depth",
  top_n           = 20,
  show_legend     = FALSE,   # suppress legend here; shared legend goes on combined figure
  base_size       = 11
)
p_raw
```

You should see that samples from the same true community have similar species
composition, with dominant species differing between communities. If
compositions look essentially uniform across all samples, communities may be
too similar (too much spillover) for the model to separate them. If so, try
reducing `spillover` in the simulation, or for real data, reconsider whether
K > 1 is warranted.

```{r plot_raw_save}
ggsave("raw_compositions.png", p_raw, width = 10, height = 4, dpi = 300)
```

---

# Fitting the model

## Basic fit

```{r fit}
fit4 <- eDNA_dmm(
  counts     = counts,
  covariates = cov_matrix,
  K          = 4,       # number of latent communities
  chains     = 4,       # labels are aligned automatically
  iter       = 4000,    # total MCMC iterations per chain (eDNA_dmm()'s own default)
  warmup     = 2000,    # warmup iterations discarded (should be ~half of iter)
  seed       = 13
)
```

4000 total iterations (2000 post-warmup per chain, 8000 pooled across 4
chains) is already generous for a dataset this size, on well-separated
ground truth like this one you should see ESS in the thousands, far past the
`> 400` "trustworthy" bar. Reach for more only if `summary()` actually reports
low ESS on a real dataset (see `vignette("eDNA_structure")`'s
*Increasing iterations for better ESS* section).

When running, Stan prints sampling progress to the console. A large block of
C++ template warnings from Eigen/Stan headers may appear; these are benign
and can be ignored.

> **Note on community labels:** a mixture likelihood does not change if the
> community labels are permuted, so chains often number the same communities
> differently. This is called label switching.
> `eDNA_dmm()` aligns every draw onto a common labelling before
> summarising, so the reported estimates and diagnostics describe the model
> rather than its naming. `fit$alignment$min_agreement` and
> `fit$alignment$lp_rhat` say whether the chains found the same communities,
> not merely the same names.

> **Note on runtime:** Stan compiles the model to C++ on the first call after
> installation. This takes 1-2 minutes and produces a lot of console output, 
> this is normal. All subsequent calls skip compilation entirely.

## Reading the output

```{r fit_print}
print(fit4)
```

```{r fit_summary}
summary(fit4)
```

`summary()` reports four things:

- **Alpha (overdispersion)**: captures compositional variance not explained by
  community structure. Typical eDNA values are 1-5. Our simulation used α = 3
  so a well-fitted model should recover something close to that.
- **Assignment certainty**: fraction of samples with >= 80% posterior probability
  of belonging to a single community. High certainty indicates well-separated
  communities; many ambiguous samples suggests K may be too high or communities
  genuinely overlap.
- **Beta coefficients**: posterior mean, 90% credible interval, and reliability
  rating for each covariate x community combination. Reliability is
  `"trustworthy"` when effective sample size (ESS) > 400.
- **HMC diagnostics**: look for zero divergent transitions and no
  `max_treedepth` warnings. See the Troubleshooting section if either appears.

## Accessing model components directly

```{r fit_components}
# Posterior community membership probabilities, one row per sample
# Columns: sample_id, prob_comm1 ... prob_commK, z_hat (MAP assignment),
#          assignment_certainty (probability of the MAP assignment)
head(fit4$sample_info)

# Posterior mean compositions: K x S matrix
# Entry [k, s] is the expected relative frequency of taxon s in community k
round(fit4$pi_mean[, 1:8], 3)

# Beta coefficient summary table
# Columns: community, covariate, mean, ci_5, ci_95, prob_positive, ess, reliability
fit4$beta_summary

# Posterior mean overdispersion: should be close to 3.0 for our simulation
fit4$alpha_mean

# Raw Stan fit object for advanced diagnostics
# E.g. extract full posterior draws for alpha:
# rstan::extract(fit4$stan_fit, pars = "alpha")$alpha
```
---

# Selecting K: how many communities?

In practice, K is rarely known in advance. The standard approach is to fit
models across a range of K values and compare them using Leave-One-Out
cross-validation (LOO-CV). LOO-CV estimates how well each model predicts
held-out samples, penalizing complexity automatically, larger K models fit
the training data better but may overfit.

`eDNA_loo()` fits the model for each K in `K_range` and compares ELPD
(expected log predictive density) across K values. Higher ELPD = better
predictive performance. Look for the "elbow" where gains flatten.

```{r loo}
loo_result <- eDNA_loo(
  counts     = counts,
  covariates = cov_matrix,
  K_range    = 2:5,
  iter       = 4000,   # eDNA_loo()'s own default; this sweep fits 4 models,
  warmup     = 2000,   # so the per-fit cost from a higher iter multiplies fast
  seed       = 13
)
```

```{r loo_plot}
# Elbow plot: the true K should show a clear gain; additional communities
# beyond K_true should produce diminishing returns
loo_result$plot
```

```{r loo_table}
# Numeric table of LOO-ELPD +/- SE for each K
loo_result$loo_table
```

```
  K       elpd       se
1 2 -8431.22  64.1
2 3 -8219.56  61.8
3 4 -8188.44  60.2   <- gain from K=3 to K=4 is large; K=4 to K=5 is negligible
4 5 -8191.03  61.5
```

**How to interpret the LOO table:**

- A large ELPD gain between adjacent K values means adding that community
  substantially improves prediction, evidence that the K-th community genuinely
  exists in the data
- When two adjacent models are within ~2 SE of each other, prefer the simpler
  (smaller K) model
- If ELPD keeps improving monotonically without an elbow, consider whether you
  have enough samples to distinguish all K communities, or whether biological
  interpretability should be the tiebreaker

> **Tip:** Some researchers also advocate fitting one model per distinct
> geographic region or sampling stratum and comparing fits alongside the LOO
> curve. This can reveal whether apparent K > 1 is driven by spatial
> heterogeneity rather than true community structure.

## The full picture: `eDNA_dmm_k_diagnostics()`

The elbow plot above is one view of "how many communities." ELPD almost
always keeps improving as K grows, a more flexible mixture fits anything
better, so on its own it rarely gives a clean stopping point.
`eDNA_dmm_k_diagnostics()` draws the fuller picture: up to six panels, side by
side, that fail in different ways on purpose. No single panel picks K; the
point is that together they usually make the choice obvious.

```{r k_diagnostics}
# Hand it loo_result directly: fits, elpd_by_run and convergence are all
# pulled out of it automatically.
k_diag <- eDNA_dmm_k_diagnostics(loo_result)

k_diag$plot     # the assembled figure
k_diag$table    # one row per K, every panel's numbers in one place
```

**What each panel is asking:**

| Panel | Question | A drop or flattening here means |
|---|---|---|
| (a) Predictive fit | LOO ELPD vs. K, averaged over chains | Rarely informative alonewith real data, because it will often steadly increase and not quite stabilize |
| (b) Marginal gain | Paired ELPD difference between adjacent K, with 2xSE | The step to this K is not worth it: filled points clear 2 SE, open points do not |
| (c) Reliability | Share of samples with a Pareto k above `pareto_threshold` (default 0.7) | The ELPD estimate at that K cannot be trusted as much ; usually the fit is unstable there anyway |
| (d) Convergence | Chain agreement after alignment (falls back to `lp__` Rhat) | The chains found genuinely different partitions at that K, more sampling will not fix this |
| (e) Assignment certainty | Distribution of each sample's largest membership probability, floor-corrected for K | Membership is spreading thin across communities instead of concentrating. Sometimes when communities are over or under parameterized, assignment certainty will drop  |
| (f) Community distinctness | Smallest Aitchison distance between any two fitted compositions | Aitchison distance is an apropriate distance metric to compositional data | If a new community is a near-copy of an existing one, the model is splitting hairs, not finding structure |

Panels (a)-(b) ask whether the model predicts better; (c)-(f) ask whether that
improvement can be trusted and is biologically meaningful. It is common, and
diagnostic, for (a) to keep climbing after (d)-(f) have already turned: that
gap is exactly a higher K buying predictive accuracy at the cost of a grouping
the data do not actually support. When all of them agree on a K, as they
should for well-separated ground truth like this example dataset, that K is
your answer with real confidence, not just the top of an elbow plot.

For datasets with more fits than convenient to hold in memory, `fits` also
accepts a function `function(k) ...` that loads each fit on demand, e.g. from
`save_dir` files written by `eDNA_loo()`. See `?eDNA_dmm_k_diagnostics` for
the full reasoning behind every panel, including the `"simple"` two-panel mode
for a quick first look.

## Extracting the best model from LOO results

The `loo_result` object stores all fitted models internally. You do not need
to refit after selecting K, extract the chosen model directly. This is
particularly important when models take hours to run.

```{r extract-best-model}
# See what K values were fitted
names(loo_result$fits)

# Review the LOO table to confirm K choice
loo_result$loo_table

# Extract the best model: no refitting needed
K_best <- 4
fit <- loo_result$fits[[paste0("K", K_best)]]

# Confirm the extracted model
print(fit)
summary(fit)
```

You can also iterate over all fits for comparative diagnostics:

```{r inspect-all-fits}
# Compare posterior mean overdispersion across K values
sapply(loo_result$fits, function(f) round(f$alpha_mean, 2))

# Compare assignment certainty: fraction of samples assigned with > 90% probability
sapply(loo_result$fits, function(f) {
  mean(f$sample_info$assignment_certainty > 0.9)
})
```

From here, `fit` is a standard `edna_dmm_fit` object and all visualization
functions work as normal.

---

# Visualization

All visualization functions return `ggplot2` objects that can be printed,
saved, or extended with additional layers. The four main visualization
functions share a consistent color palette so that the same community always
appears in the same color across all plots; this consistency is essential for
readable multi-panel figures.

## Raw species compositions: `plot_true_compositions()`

Already covered in section 4. Key arguments for reference:

```{r plot_raw_options}
# Faceted by a variable, sorted within panels, top 20 taxa shown
plot_true_compositions(
  counts,
  metadata        = sim$metadata,
  facet_var       = "TrueCommunity",   # facet columns by this variable
  facet_row_var   = NULL,              # facet rows by this variable (e.g. "Year")
  sort_var        = "Depth",           # sort samples within panels by this variable
  top_n           = 20,                # top N taxa shown; rest collapsed to "Other"
  show_legend     = TRUE,
  legend_nrow     = 8,                 # the legend has room for 8 rows
  legend_position = "bottom",
  base_size       = 11,
  vline_var       = NULL,              # draw vertical reference line at a covariate value
  vline_value     = NULL               # e.g. vline_var = "Latitude", vline_value = 40.44
)
```

## Community assignment: `eDNA_dmm_structure()`

The structure plot shows one vertical bar per sample, divided into colored
segments whose heights equal the posterior probability of belonging to each
community. This is the primary diagnostic for how well the model separated
communities.

```{r struct_basic}
# Minimal call: all samples, single panel, no sorting
eDNA_dmm_structure(fit)
```

```{r struct_faceted}
# Faceted by true community, sorted by depth within each panel
p_struct <- eDNA_dmm_structure(
  fit,
  metadata        = sim$metadata,
  facet_var       = "TrueCommunity",
  sort_var        = "Depth",
  show_legend     = FALSE,   # suppress here; add to combined figure
  base_size       = 11
)
p_struct
```

With well-separated communities, each panel should be dominated by one color.
Mixed bars indicate either genuinely ambiguous samples (samples at community
boundaries) or label switching (communities are recovered but numbered
differently from the true labels, a permutation, not an error).

`eDNA_dmm()` already ran `dmm_align_labels()` for you before `fit` was ever
returned, so cross-chain label switching should
already be resolved, mixed bars here are more likely genuine ambiguity. The
one residual case alignment cannot always catch on its own is a single chain
relabelling *mid-run* (rare, but see `?dmm_align_labels`'s *Why any fit may
need this* section for when it happens): check `fit$alignment$pct_permuted`
to see how much relabelling was needed in the first place, and if the
structure plot still looks wrong where the diagnostics look clean, try
re-aligning with a different algorithm:

```{r struct_realign, eval=FALSE}
fit_ecr <- dmm_align_labels(fit, method = "ECR-pivot")
eDNA_dmm_structure(fit_ecr, metadata = sim$metadata, facet_var = "TrueCommunity")
```

```{r struct_row_col_facet}
# For real data with two grouping variables (e.g., depth stratum x year)
# facet_row_var creates rows of panels; facet_var creates columns
# This is the recommended layout for publication figures with two-way structure
eDNA_dmm_structure(
  fit,
  metadata        = sim$metadata,
  facet_var       = "TrueCommunity",   # columns
  facet_row_var   = NULL,              # rows: set to e.g. "DepthStratum"
  sort_var        = "Depth",
  vline_var       = NULL,              # draw a vertical reference line
  vline_value     = NULL               # e.g. biogeographic boundary latitude
)
```

```{r struct_custom_colors}
# Custom colors: useful for matching a manuscript color scheme
p_struct_colors <- eDNA_dmm_structure(
  fit,
  metadata         = sim$metadata,
  facet_var        = "TrueCommunity",
  sort_var         = "Depth",
  community_colors = c(
    "Community 1" = "#E63946",
    "Community 2" = "#457B9D",
    "Community 3" = "#2A9D8F",
    "Community 4" = "#E9C46A"
  )
)
p_struct_colors
```

```{r struct_ggplot_extend}
# ggplot2 layers can be added directly to the returned object
p_struct_colors +
  theme(strip.text = element_text(size = 13, face = "bold")) +
  labs(caption = "Panels = true community | Sorted by depth within panel")
```

```{r struct_save}
ggsave("structure_plot.png", p_struct_colors, width = 10, height = 4, dpi = 300)
```

## Publication layouts: rows, labels and legends

`plot_true_compositions()` and `eDNA_dmm_structure()` share one layout engine, so
the two can be stacked into a single figure whose panels line up exactly. This
section covers the arguments that matter once a plot stops being a quick look and
starts being a manuscript figure.

The examples below need a second grouping variable to demonstrate two-way
faceting (`facet_var` for columns, `facet_row_var` for rows). Our simulated
data only has one categorical variable of its own, `TrueCommunity`, so `meta`
adds a second, `depth_bin`, the way you would with any real second grouping
(site, year, depth stratum, ...):

```{r layout_meta}
meta <- sim$metadata
meta$depth_bin <- ifelse(meta$Depth > median(meta$Depth), "deep", "shallow")
```

### Two-way panels and why they are built with cowplot

`facet_var` makes panel **columns**; `facet_row_var` makes panel **rows**. When
`facet_row_var` is supplied, one sub-plot is built per row level and the sub-plots
are stacked with cowplot, rather than using a single `facet_grid()`.

That is deliberate. It is the only arrangement that keeps panel widths
proportional to the number of samples in each panel (via `space = "free_x"`), so a
stratum with 30 samples is drawn three times wider than one with 10 instead of
both being stretched to equal width.

```{r layout_two_way, eval=FALSE}
eDNA_dmm_structure(
  fit,
  metadata      = meta,
  facet_var     = "TrueCommunity",   # columns (use your own second grouping,
  facet_row_var = "depth_bin",       # rows     e.g. "year" or "site", for real data)
  sort_var      = "Depth"
)
```

There is one consequence worth knowing before you build a figure around it:

```{r layout_cowplot_gotcha, eval=FALSE}
# WITHOUT facet_row_var -> a ggplot object, so ggplot2 layers can be added:
p <- eDNA_dmm_structure(fit, metadata = meta)
p + labs(caption = "works fine")

# WITH facet_row_var -> a cowplot grid object. `+ theme(...)` will NOT work.
# Style it through the function's own arguments instead (base_size,
# strip_text_size, axis_text_size, row_label_size, ylab_size ...).
p2 <- eDNA_dmm_structure(fit, metadata = meta, facet_row_var = "depth_bin")
# p2 + labs(...)   # <- error
```

### Row labels that generalize: `row_label_fn`

Each row panel gets a label derived from its `facet_row_var` level. `row_label_fn`
is the formatter applied to that level, which is what lets the same code produce
sensible labels for depth, year, site or season:

```{r layout_row_labels, eval=FALSE}
# depth_bin values 10, 50, 200 -> "10 m", "50 m", "200 m"
row_label_fn = function(x) paste0(x, " m")

# year values 2021, 2022 -> "Year: 2021", "Year: 2022"
row_label_fn = function(x) paste0("Year: ", x)
```

`row_label_position` decides where that label is drawn:

- `"title"` (default), a bold label above each row's panel.
- `"ylab"`, folded into the y-axis label, which was the older, hardcoded behavior.

With `"title"`, the y-axis label is drawn once for the whole stack and is taken
from `ylab` (`"Proportion"` for `plot_true_compositions()`,
`"Membership probability"` for `eDNA_dmm_structure()`).

### Legends that fit

Taxon legends get large quickly. `legend_nrow` / `legend_ncol` set the shape,
`legend_key_size` and `legend_text_size` set the scale, and `legend_rel_height` /
`legend_rel_width` decide how much of the figure the legend is allowed to occupy.

```{r layout_legend, eval=FALSE}
plot_true_compositions(
  counts,
  metadata          = meta,
  facet_row_var     = "depth_bin",
  show_legend       = TRUE,
  legend_position   = "bottom",
  legend_ncol       = 7,      # one colour family per column (see n_hues)
  legend_key_size   = 0.35,
  legend_text_size  = 7,
  legend_rel_height = 0.22    # give the legend room: see the warning below
)
```

Taxa are laid out hue block by hue block, `n_hues` families in total (default 7),
so setting `legend_ncol = n_hues` puts roughly one colour family per column and
makes a 20-taxon legend scannable.

> **Watch out:** `legend_rel_height` is a hard allocation, not a minimum. The
> legend receives exactly that fraction of the figure and any rows that do not fit
> are **clipped silently**. If legend entries seem to have vanished, raise
> `legend_rel_height`, the taxa are still in the plot.

### Nesting a plot inside a bigger figure: `panel_labels`

By default each row panel is lettered (a, b, c...). When the plot is itself one
panel of a larger composite, turn that off so the inner letters do not collide
with the outer ones:

```{r layout_panel_labels, eval=FALSE}
eDNA_dmm_structure(fit, metadata = meta, facet_row_var = "depth_bin",
                   panel_labels = FALSE)
```

### Making both plots share one taxon palette: `taxa_include`

The taxon palette is a deterministic function of the sorted taxon set: the same
set in gives the same colours out. So if the observed plot picks its own top 20
and the fitted plot independently picks its own, the two sets differ and any
shared taxon is drawn in **two different colours**. `dmm_taxon_order()` exists to
prevent that, rank once, pass the same set to both:

```{r layout_shared_palette, eval=FALSE}
taxa <- dmm_taxon_order(fit, top_n = 20)

common <- list(
  metadata      = meta,
  facet_var     = "TrueCommunity",
  facet_row_var = "depth_bin",
  sort_var      = "Depth",
  row_label_fn  = function(x) paste0(x, " m"),
  base_size     = 11
)

p_obs <- do.call(plot_true_compositions,
                 c(list(counts, taxa_include = taxa), common))
p_fit <- do.call(eDNA_dmm_structure,
                 c(list(fit, panel_labels = FALSE), common))

cowplot::plot_grid(p_obs, p_fit, ncol = 1, rel_heights = c(1, 1))
```

See *Multi-panel publication figures* below for assembling and saving the
finished figure - in particular, use `cowplot::save_plot()` rather than
`ggsave()` for anything built with `plot_grid()`.

Note that `dmm_taxon_order()` ranks taxa by total posterior weight across
communities (`colSums(pi_mean)`), *not* by mean observed frequency. A taxon that
dominates a single community ranks highly even when it is rare overall, which is
the ordering you want when the figure's job is to show what distinguishes the
communities from one another.

### Reference lines: `vline_*`

Draws a vertical line where a numeric covariate crosses a threshold, once samples
have been sorted along the x-axis:

```{r layout_vline, eval=FALSE}
plot_true_compositions(
  counts,
  metadata        = meta,
  sort_var        = "Depth",
  vline_var       = "Depth",   # column the threshold is measured in
  vline_value     = 200,       # the threshold itself
  vline_linetype  = "dashed",
  vline_linewidth = 0.7
)
```

## Posterior community compositions: `eDNA_dmm_compositions()`

`eDNA_dmm_compositions()` shows what each latent community looks like in
species space, the model's posterior mean estimate of each community's
taxonomic signature. This is the complement of the structure plot: the
structure plot shows which community each *sample* belongs to; the compositions
plot shows which *species* characterize each community.

Colors match those used in `plot_true_compositions()` and
`eDNA_dmm_structure()`, so the same species always appears in the same color
across all three plots.

```{r compositions_basic}
# Basic: top 20 taxa, legend on right, x-axis shows community numbers (1, 2, 3...)
p_comp <- eDNA_dmm_compositions(
  fit,
  top_n           = 20,
  base_size       = 13,
  legend_position = "right"
)
p_comp
```

**What to look for:**

- Each community bar should be dominated by a distinct set of species
- Species shared between communities appear in multiple bars but at different
  relative heights; this reflects the spillover structure baked into the simulation
- Compare with `plot_true_compositions()` faceted by TrueCommunity: the species
  pattern in each community bar should mirror the raw data for samples assigned
  to that community
- If two community bars look nearly identical, those communities may not be
  genuinely distinct in species space, consider reducing K

```{r compositions_save}
ggsave("community_compositions.png", p_comp, width = 8, height = 5, dpi = 300)
```

## Multi-panel publication figures

Because `plot_true_compositions()`, `eDNA_dmm_structure()`, and
`eDNA_dmm_compositions()` all use the same color palette and cowplot-based
architecture, they align precisely when stacked. The panels are designed to
tell a coherent story: what the data look like (A), how the model assigned
samples (B), and what each community looks like in species space (C).

> **Important:** Use `cowplot::save_plot()` rather than `ggsave()` when saving
> figures assembled with `plot_grid()`. `save_plot` correctly handles the
> multi-panel dimensions and prevents panels from being squished.

### Two-panel: raw vs. assigned (A/B figure)

The canonical before/after figure for a publication: panel A shows observed
species frequencies, panel B shows the model's community assignments for the
same samples in the same order. The two panels share a layout so visual
comparison is immediate.

```{r cowplot_ab, eval=FALSE}
p_raw <- plot_true_compositions(
  counts,
  metadata        = sim$metadata,
  facet_var       = "TrueCommunity",
  sort_var        = "Depth",
  show_legend     = FALSE     # shared legend goes at the bottom of the combined figure
)

p_struct <- eDNA_dmm_structure(
  fit,
  metadata        = sim$metadata,
  facet_var       = "TrueCommunity",
  sort_var        = "Depth",
  show_legend     = TRUE,
  legend_position = "bottom"
)

fig_ab <- cowplot::plot_grid(
  p_raw, p_struct,
  ncol        = 1,
  labels      = c("A", "B"),
  label_size  = 16,
  rel_heights = c(1, 1)
)
print(fig_ab)

cowplot::save_plot(
  "fig_raw_vs_assigned.png",
  fig_ab,
  base_width  = 10,
  base_height = 8
)
```

**What to look for:** In a well-fitted model, the color pattern in panel B
should mirror the species pattern in panel A, samples that share a dominant
species signal in A should share a dominant community color in B. If panel B
is mostly mixed bars where panel A shows a clear signal, try more MCMC
iterations or a different K.

### Three-panel: raw -> assigned -> community signatures (A/B/C figure)

Adding panel C creates a complete narrative: A shows the raw data, B shows
which community each sample belongs to, and C shows what each of those
communities looks like in species space.

```{r cowplot_abc, eval=FALSE}
p_raw <- plot_true_compositions(
  counts,
  metadata        = sim$metadata,
  facet_var       = "TrueCommunity",
  sort_var        = "Depth",
  show_legend     = FALSE
)

p_struct <- eDNA_dmm_structure(
  fit,
  metadata        = sim$metadata,
  facet_var       = "TrueCommunity",
  sort_var        = "Depth",
  show_legend     = FALSE
)

p_comp <- eDNA_dmm_compositions(
  fit,
  top_n           = 20,
  legend_position = "right"
)

fig_abc <- cowplot::plot_grid(
  p_raw, p_struct, p_comp,
  ncol        = 1,
  labels      = c("A", "B", "C"),
  label_size  = 16,
  rel_heights = c(1, 1, 1)
)
print(fig_abc)

cowplot::save_plot(
  "fig_three_panel.png",
  fig_abc,
  base_width  = 10,
  base_height = 12
)
```

> **Tip for real data with `facet_row_var`:** When using depth stratum x year
> panels, the A/B panels can become very tall. In that case, assemble A and B
> as one figure and C as a separate supplementary figure, or place A and B
> side-by-side with `ncol = 2` and C below.

## NMDS ordination: `eDNA_dmm_nmds()`

NMDS provides a complementary view of community structure: samples are
positioned in low-dimensional space according to their compositional
dissimilarity, then colored by their MAP community assignment. This makes it
easy to see how well communities separate in multivariate space independently
of the model.

For metabarcoding data the eDNA index (Kelly et al. 2019) is applied before
computing Bray-Curtis distances to account for compositional structure.
Alternatively, Jaccard distance operates on presence/absence and is more
conservative.

```{r nmds_basic}
nmds_result <- eDNA_dmm_nmds(fit)
nmds_result$plot

# Always check the stress value
message("NMDS stress: ", round(nmds_result$nmds$stress, 3))
# < 0.05 = excellent; 0.05-0.10 = good; 0.10-0.20 = adequate; > 0.20 = poor
# If stress > 0.20, switch to k = 3 and display two axis pairs
```

```{r nmds_options}
# 3D NMDS for lower stress; display axes 1 and 3
nmds_3d <- eDNA_dmm_nmds(fit, k = 3, nmds_axes = c(1, 3))
nmds_3d$plot

# Jaccard (presence/absence): more conservative, ignores abundance information
eDNA_dmm_nmds(fit, distance = "jaccard", show_ellipse = FALSE)$plot

# Custom styling: larger point size range makes certainty differences clearer
eDNA_dmm_nmds(
  fit,
  show_ellipse     = TRUE,
  ellipse_type     = "t",       # "t" = robust (recommended); "norm" = normal-based
  size_range       = c(2, 6),
  alpha            = 0.85,
  community_colors = c(
    "Community 1" = "#E63946",
    "Community 2" = "#457B9D",
    "Community 3" = "#2A9D8F",
    "Community 4" = "#E9C46A"
  )
)$plot
```

```{r nmds_save}
ggsave("nmds.png", nmds_result$plot, width = 7, height = 6, dpi = 300)
```

## Covariate effects: `eDNA_dmm_beta()`

The beta plot overlays prior and posterior distributions for each softmax
regression coefficient. A posterior pulled away from the prior (the grey
density) is evidence that the covariate genuinely predicts community
membership. A posterior that closely matches the prior means the data were
uninformative about that effect, a valid scientific result, not a failure.

```{r beta_joint}
# Joint layout (default): all communities overlaid per covariate panel
# Best for K <= 4; easy to see which communities are pulled in which direction
beta_result <- eDNA_dmm_beta(fit)
beta_result$plot
```

```{r beta_separate}
# Separate layout: one row per community, one column per covariate
# Better for K > 4 where community densities overlap too much in joint layout
eDNA_dmm_beta(fit, layout = "separate")$plot
```

```{r beta_table}
# Posterior summary table with directional probabilities and ESS
beta_result$table
# Columns: community, covariate, mean, ci_5, ci_95,
#          prob_positive, prob_negative, ess, reliability
```

### Covariate effects as intervals: `eDNA_dmm_beta_intervals()`

`eDNA_dmm_beta()`'s densities show how far the data moved a coefficient from
its prior, but reading a density's tail against zero is a squint. This is the
same coefficients as a coefficient (dot-and-line) plot instead: one row per
community x covariate, a point estimate, two nested credible intervals, and a
dashed line at zero, so "does this cross zero" is immediate.

```{r beta_intervals_joint}
beta_intervals <- eDNA_dmm_beta_intervals(fit)
beta_intervals$plot
```

A **filled** point means the outer interval (90% by default) excludes zero, a
real effect at that credibility level; a **hollow** point means it crosses
zero. The thick line is the inner interval (50%), the thin line the outer one.

```{r beta_intervals_options}
# Narrower/wider intervals than the 50%/90% default
eDNA_dmm_beta_intervals(fit, intervals = c(0.5, 0.95))$plot

# Compare magnitudes across covariates on one shared x scale, instead of each
# covariate getting its own range (the default, "free_x")
eDNA_dmm_beta_intervals(fit, facet_scales = "fixed")$plot

# The table behind the plot: one row per coefficient
beta_intervals$table
# Columns: community, covariate, estimate, the four interval bounds, excludes_zero
```

Both `eDNA_dmm_beta()` and `eDNA_dmm_beta_intervals()` take the same
`reference` argument (see *Interpreting the beta plot* just below), so
whichever reading you pick, contrast-against-one-community or
deviation-from-average, is consistent across both plot types.

**Interpreting the beta plot for our simulation:**

Our design has four communities defined by depth x shore distance:

| Community | Depth | Shore distance | Expected beta (Depth) | Expected beta (Distance) |
|-----------|-------|----------------|-----------------------|--------------------------|
| 1 | Deep | Offshore | positive | positive |
| 2 | Shallow | Offshore | negative | positive |
| 3 | Deep | Inshore | positive | negative |
| 4 | Shallow | Inshore | negative | negative |

`eDNA_dmm()` uses a **centred** softmax parameterization: there is no reference
community. For each covariate the four communities' coefficients are
constrained to sum to zero, so each one is that community's *deviation from
the average community* on that covariate, not a contrast against a fixed
baseline. Community 1 and 3 sit above the average Depth (70 vs. the grand mean
of 55), so both get a positive Depth coefficient; Community 2 and 4 sit below
it, so both get a negative one. The same logic applies to Distance_shore
against its grand mean (120). Posteriors pulled in the expected directions
confirm parameter recovery. (To read coefficients as a contrast against one
particular community instead, subtract that community's row from the
posterior draws.)

```{r beta_options}
# Plot a single covariate only
eDNA_dmm_beta(fit, covariates_to_plot = "Depth")$plot

# Include the intercept (community prevalence in covariate-free space)
eDNA_dmm_beta(fit, show_intercept = TRUE)$plot

# If you modified the prior SD in Stan, match it here for correct visualization
eDNA_dmm_beta(fit, beta_prior_sd = 2.5)$plot
```

```{r beta_save}
ggsave("beta_plot.png", beta_result$plot, width = 9, height = 5, dpi = 300)
```

---

# Verifying recovery of true parameters

One of the main advantages of working with simulated data is that you can
compare fitted values to the known ground truth. This section shows three
verification checks.

## Community assignments

```{r verify_assignments}
# Compare MAP community assignment to the true community label
comparison <- data.frame(
  sample_id   = fit$sample_info$sample_id,
  fitted_comm = fit$sample_info$z_hat,
  true_comm   = paste0("Community ", sim$metadata$TrueCommunity),
  certainty   = round(fit$sample_info$assignment_certainty, 2)
)
print(comparison)

# Cross-tabulation: perfect recovery gives a permutation matrix
# (each true community mapped to exactly one fitted community, possibly reordered)
table(comparison$fitted_comm, comparison$true_comm)
```

Note that label switching means "Community 1" in the fit may not correspond to
"Community 1" in the simulation. What matters is the 1:1 mapping: each row and
column of the cross-tabulation should have exactly one non-zero entry.

## Alpha (overdispersion) recovery

```{r verify_alpha}
message("True alpha:   3.0")
message("Fitted alpha: ", round(fit$alpha_mean, 2))

# Visualize the full posterior for alpha
alpha_draws <- rstan::extract(fit$stan_fit, pars = "alpha")$alpha
hist(alpha_draws, breaks = 40,
     main = "Posterior distribution for overdispersion (alpha)",
     xlab = "alpha", col = "steelblue", border = "white")
abline(v = 3.0, col = "firebrick", lwd = 2, lty = 2)
legend("topright", legend = "True value (alpha = 3)",
       col = "firebrick", lty = 2, lwd = 2, bty = "n")
```

## Beta coefficient recovery

```{r verify_beta}
# Check the beta table against the expected directions from the table above
beta_result$table[, c("community", "covariate", "mean", "ci_5", "ci_95",
                       "prob_positive", "reliability")]
```


---

# Working with real data

The workflow for real eDNA metabarcoding data is identical to the simulation
workflow; you just skip the simulation steps.

```{r real_data_workflow, eval=FALSE}
# 1. Load your count matrix (samples x taxa, integer read counts)
counts <- read.csv("my_count_table.csv", row.names = 1) |> as.matrix()

# 2. Load metadata / environmental covariates
metadata <- read.csv("my_metadata.csv")

# 3. Align them: this step is critical and easy to get wrong
counts   <- counts[order(rownames(counts)), ]
metadata <- metadata[order(metadata$sample_id), ]
stopifnot(all(rownames(counts) == metadata$sample_id))

# 4. Inspect raw compositions before fitting
plot_true_compositions(
  counts,
  metadata  = metadata,
  facet_var = "site",       # or "season", "depth_stratum", etc.
  sort_var  = "depth",
  top_n     = 20
)

# 5. Select covariates: use continuous, Z-score-compatible variables
#    Year as a factor often violates linearity; test it post-hoc if needed
cov_matrix <- metadata[, c("depth", "latitude")]

# 6. Select K using LOO cross-validation
loo_result <- eDNA_loo(counts, cov_matrix, K_range = 2:6)
loo_result$plot
loo_result$loo_table

# 7. Extract the chosen K from LOO results: no refitting needed
K_best <- 4   # update based on LOO elbow
fit <- loo_result$fits[[paste0("K", K_best)]]

# 8. Visualize
p_raw    <- plot_true_compositions(counts, metadata = metadata,
                                   facet_var = "site", sort_var = "depth")
p_struct <- eDNA_dmm_structure(fit, metadata = metadata,
                               facet_var = "site", sort_var = "depth")
p_comp   <- eDNA_dmm_compositions(fit, top_n = 20)
p_nmds   <- eDNA_dmm_nmds(fit)$plot
p_beta   <- eDNA_dmm_beta(fit)$plot

# 9. Assemble publication figure
fig <- cowplot::plot_grid(
  p_raw, p_struct,
  ncol       = 1,
  labels     = c("A", "B"),
  label_size = 16
)
cowplot::save_plot("fig_main.png", fig, base_width = 10, base_height = 8)
```

---

# Troubleshooting

## Divergent transitions

```
Warning: There were X divergent transitions after warmup.
```

The sampler struggled with the posterior geometry. Try in order:

1. Increase `adapt_delta` toward `0.99`
2. Increase `max_treedepth` to 14 or 15
3. Reduce K, more communities create more complex posteriors
4. Check for all-zero rows/columns in your count matrix

A small number of divergences (< 1% of iterations) in a large run can sometimes
be tolerated, but results should be interpreted with caution. Many divergences
indicate a fundamental geometry problem.

## Low ESS for beta parameters

ESS < 100 for beta coefficients usually means the covariate has little effect
on community membership. If the posterior closely overlays the prior in the
beta plot, the data are uninformative about that covariate effect. This is a
valid scientific result, not a modeling failure. If ESS is low but the
posterior is clearly pulled from the prior, try increasing `iter` and `warmup`.

## Very long runtime

Each iteration requires N x K evaluations of the Dirichlet-Multinomial
likelihood, which scales with sample size and K. Practical strategies:

- Use `iter = 2000, warmup = 1000` for exploratory fits, then increase for
  final inference
- Filter to the top 50-100 taxa by total read count to reduce S without losing
  the dominant signal
- For LOO with K_range = 2:6, plan for roughly 20-40 minutes per K value on a
  modern laptop

## Function not found after installation

Make sure you are using the latest version of the package:

```{r reinstall, eval=FALSE}
detach("package:eDNAstructure", unload = TRUE)
remove.packages("eDNAstructure")
remotes::install_github("pedrobdfp/eDNA_structure", upgrade = "never")
library(eDNAstructure)
```

---

# Session information

```{r session}
sessionInfo()
```
