---
title: "eDNA_structure: Full Tutorial"
subtitle: "From Simulation to Fitted Communities — A Complete Walkthrough"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 4
    number_sections: true
vignette: >
  %\VignetteIndexEntry{eDNA_structure: Full Tutorial}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse   = TRUE,
  comment    = "#>",
  eval       = FALSE,
  fig.width  = 8,
  fig.height = 5,
  message    = FALSE,
  warning    = FALSE
)
```

```{r load}
library(eDNAstructure)
library(ggplot2)   # for post-hoc plot customization
```

---

# Overview

This tutorial walks through the complete `eDNAstructure` workflow:

1. **Simulate** a realistic eDNA survey with known ground truth
2. **Inspect** the raw community signal before fitting
3. **Format** data correctly for `eDNA_dmm()`
4. **Select K** using LOO cross-validation
5. **Fit** the Dirichlet-Multinomial Mixture model
6. **Visualize** community assignments, NMDS ordination, and covariate effects
7. **Verify** recovery of true parameters

For a shorter introduction, see `vignette("eDNA_structure")`.

---

# The simulation pipeline

`eDNAstructure` includes a full mechanistic simulation pipeline so you can
generate datasets with known community structure and verify that the model
recovers it. The same pipeline was used to generate the built-in example
dataset (`get_example_data()`).

## Why simulate?

- **Controlled validation**: you know the true K, true compositions, and true
  covariate effects — so you can check that the model gives the right answers
- **Power analysis**: how many samples do you need? What spillover level is too
  high for the model to separate communities?

## The five simulation steps

The pipeline runs five sequential steps. You can call them individually for
full control, or use `simulate_eDNA_survey()` to run all five at once.

### Step 1 — Community compositions

`generate_community_compositions()` builds a K × S matrix where each row is
one community's species frequency vector.

```{r sim_compositions}
# Four communities, 40 species
# Communities 1+3 share a group (deep habitat); communities 2+4 share a group (surface)
community_mat <- generate_community_compositions(
  n_communities         = 4,
  n_species             = 40,
  n_dominant_range      = c(2, 3),   # 2-3 dominant species per community
  n_unique_low_freq     = 4,         # 4 species unique to each community
  community_groups      = c(1, 2, 1, 2),  # depth groups
  n_group_shared        = 4,         # 4 species shared within each depth group
  group_shared_presence = 1.0,
  dominant_freq         = c(0.08, 0.14),
  low_unique_freq       = c(0.02, 0.04),
  group_shared_freq     = c(0.02, 0.04),
  shared_freq           = c(0.02, 0.04),
  shared_presence       = 0,         # no globally shared species
  spillover             = 0.15,      # 15% of a species' frequency leaks to other communities
  seed                  = 2026
)

dim(community_mat)    # 4 x 40
round(rowSums(community_mat), 6)   # each row sums to 1
```

The `spillover` parameter is key: `0` gives perfectly exclusive communities
(each species present in exactly one community); `0.15` gives realistic partial
overlap. Higher values make communities harder to distinguish.

### Step 2 — Contributors (organisms)

`generate_contributors()` assigns individual organisms to each sample, drawing
species from the community frequency vector.

```{r sim_contributors}
contrib_obj <- generate_contributors(
  community_compositions = community_mat,
  samples_per_community  = 5,          # 5 sampling stations per community = 20 total
  n_contributors_range   = c(20, 100), # 20-100 organisms per sample
  n_distribution         = "Negative Binomial",  # overdispersed, more realistic
  size_range             = c(1, 10),   # body size range (affects eDNA shedding)
  distance_range         = c(0, 100),  # distance from sampler (affects eDNA decay)
  seed                   = 2026
)

length(contrib_obj$contributors_list)  # 20 samples
head(contrib_obj$contributors_list[[1]])
```

### Step 3 — eDNA shedding and sub-sampling

`generate_eDNA()` models how eDNA molecules move from organisms into the water
column and are captured by a sampling bottle.

```{r sim_edna}
eDNA_obj <- generate_eDNA(
  contributors_list = contrib_obj$contributors_list,
  shedding_rate     = 1000,    # baseline molecules shed per unit size
  beta              = 0.75,    # allometric exponent (metabolic scaling)
  decay_rate        = 0.1,     # exponential distance decay
  shedding_error    = 0.3,     # lognormal noise on shedding (SD on log scale)
  bottle_volume     = 0.1,     # fraction of local pool captured
  bio_reps          = 1        # one bottle per station
)

head(eDNA_obj$Summary)
```

### Step 4 — Metabarcoding

`simulate_metabarcoding()` applies amplification bias and sequencing noise to
produce final integer read counts.

```{r sim_metab}
metab_df <- simulate_metabarcoding(
  eDNA_obj,
  mean_read_depth = 10000,    # mean reads per sample
  read_depth_sd   = 0.2,      # lognormal SD for read depth variation
  error_sd        = 0.05,     # Gaussian noise on species frequencies
  rep             = 1         # one sequencing replicate per bottle
)

str(metab_df)
```

### Step 5 — Environmental covariates

`generate_sample_covariates()` assigns realistic covariate values to each sample
drawn from community-specific distributions.

```{r sim_covariates}
# Define where each community lives in covariate space
community_covariate_means <- matrix(
  c(80, 200,   # Community 1: deep + offshore
    10, 200,   # Community 2: surface + offshore
    80,  20,   # Community 3: deep + inshore
    10,  20),  # Community 4: surface + inshore
  nrow = 4, byrow = TRUE,
  dimnames = list(NULL, c("Depth", "Distance_shore"))
)

sample_metadata <- generate_sample_covariates(
  contributors_list    = contrib_obj$contributors_list,
  community_covariates = community_covariate_means,
  covariate_sds        = NULL   # default SD = 5 for all
)

head(sample_metadata)
```

## Running the full pipeline at once

For most uses, `simulate_eDNA_survey()` runs all five steps in one call and
returns a count matrix and covariate data frame ready for `eDNA_dmm()`.

```{r sim_full}
sim <- simulate_eDNA_survey(
  n_communities             = 4,
  n_species                 = 40,
  samples_per_community     = 5,
  community_covariate_means = community_covariate_means,
  mean_read_depth           = 10000,
  spillover                 = 0.15,
  shedding_error            = 0.3,
  seed                      = 2026
)

# Ready-to-use outputs
dim(sim$counts)        # 20 samples x 40 taxa
head(sim$covariates)   # sample_id, TrueCommunity, Depth, Distance_shore
```

---

# Data format

Before fitting, make sure your data are in the right format. This section
explains the required format using the simulated data as an example.

## The count matrix

The count matrix is the primary input to `eDNA_dmm()`.

```{r data_format_counts}
# Rows = samples, columns = taxa, values = integer read counts
sim$counts[1:4, 1:6]

# Each row should have a meaningful total — check
range(rowSums(sim$counts))
hist(rowSums(sim$counts), main = "Read depth distribution", xlab = "Total reads")

# Zero-count columns are dropped automatically with a warning
# but it's good practice to remove them yourself
counts <- sim$counts[, colSums(sim$counts) > 0]
dim(counts)
```

**Key rules:**
- All values must be non-negative integers (or coercible to integers)
- No NA values (replace with 0 if a taxon was simply absent)
- No all-zero rows (remove samples with zero total reads)
- Row names are used as sample IDs in plots — set them meaningfully

## The covariate data frame

```{r data_format_covs}
# Same row order as counts — this is critical
head(sim$covariates)

# Verify alignment
stopifnot(all(rownames(counts) == sim$covariates$sample_id))

# Select only the covariates you want to model
cov_matrix <- sim$covariates[, c("Depth", "Distance_shore")]
head(cov_matrix)
```

Covariates are Z-score standardized internally when `scale_covariates = TRUE`
(the default). Do not pre-standardize — let the function handle it so
beta coefficients remain on a common scale.

---

# Visualizing raw compositions (before fitting)

Before running the model, it is useful to visualize the raw species composition
of each sample. `plot_true_compositions()` produces a stacked bar plot in the
same format as the structure plot, making it easy to compare the two side by side.

```{r plot_raw}
# All samples, no faceting
plot_true_compositions(counts)
```

```{r plot_raw_faceted}
# Faceted by true community — this is only possible with simulated data where
# the true community labels are known. With real data, facet by a known
# grouping variable like depth stratum or site.
plot_true_compositions(
  counts,
  metadata  = sim$covariates,
  facet_var = "TrueCommunity",
  sort_var  = "Depth",
  top_n     = 20
)
```

You should see that samples from the same true community have similar species
composition, and that compositions differ between communities. If you cannot see
any structure at all, the communities may be too similar (too much spillover)
for the model to separate them.

```{r plot_raw_save}
p_raw <- plot_true_compositions(
  counts,
  metadata  = sim$covariates,
  facet_var = "TrueCommunity"
)
#ggsave("raw_compositions.png", p_raw, width = 10, height = 4, dpi = 300)
```

---

# Selecting K: how many communities?

Rarely do you know K in advance. The standard approach is to fit models for
K = 2, 3, 4, 5 and compare them using Leave-One-Out cross-validation (LOO-CV).

`eDNA_loo()` does this automatically. It takes a few minutes per K value.

```{r loo, cache=TRUE}
loo_result <- eDNA_loo(
  counts     = counts,
  covariates = cov_matrix,
  K_range    = 2:5,
  iter       = 4000,
  warmup     = 2000,
  seed       = 13
)
```

```{r loo_plot}
# Elbow plot — look for the K where gains flatten
loo_result$plot
```

```{r loo_table}
# Numeric table of LOO-ELPD ± SE
loo_result$loo_table
```

```
  K       elpd       se
1 2 -8431.22  64.1
2 3 -8219.56  61.8
3 4 -8188.44  60.2   ← gain flattens here
4 5 -8191.03  61.5
```

**How to read this:**

- Higher LOO-ELPD = better predictive performance
- The gain from K=3 to K=4 should be large (the true K); from K=4 to K=5
  it should be small (diminishing returns)
- If two adjacent models are within ~2 SE of each other, prefer the simpler one

For data generated with K=4, the elbow should appear at K=4.

---

# Fitting the model

## Basic fit

```{r fit, cache=TRUE}
fit <- eDNA_dmm(
  counts     = counts,
  covariates = cov_matrix,
  K          = 4
)
```

When you run this, Stan prints sampling progress to the console. This is normal.
The first run compiles the Stan model (1–2 minutes), after which subsequent runs
skip straight to sampling.

## Reading the output

```{r fit_print}
print(fit)
```

```
eDNA Dirichlet-Multinomial Mixture Model
=========================================
  K (communities)  : 4
  N (samples)      : 20
  S (taxa)         : 40
  Covariates       : Depth, Distance_shore
  Mean alpha       : 3.14

Community sizes (MAP assignment):
  Community 1     : 5 samples (25.0%)
  Community 2     : 5 samples (25.0%)
  Community 3     : 5 samples (25.0%)
  Community 4     : 5 samples (25.0%)
```

```{r fit_summary}
summary(fit)
```

`summary()` reports:

- **Alpha**: overdispersion. Typical eDNA values are 1–5. Our simulated data
  used alpha = 3, so the model should recover something close to that.
- **Assignment certainty**: fraction of samples with ≥80% posterior probability
  of belonging to one community. High certainty = well-separated communities.
- **Beta coefficients**: effect of depth and distance from shore on community
  membership. Reliability is `"trustworthy"` when ESS > 400.
- **HMC diagnostics**: look for zero divergences and no max treedepth warnings.

## Accessing model components

```{r fit_components}
# Posterior membership probabilities (one row per sample)
head(fit$sample_info)
# Columns: sample_id, prob_comm1, prob_comm2, prob_comm3, prob_comm4,
#          z_hat (MAP assignment), assignment_certainty

# Posterior mean compositions (K x S)
round(fit$pi_mean[, 1:8], 3)

# Beta coefficient table
fit$beta_summary

# Posterior mean overdispersion
fit$alpha_mean   # should be close to 3.0 (the simulated truth)
```

---

# Visualization

All three visualization functions return `ggplot2` objects. You can print them,
save them with `ggsave()`, or modify them with additional ggplot2 layers.

## Structure plot

```{r struct_basic}
# Basic: all samples, single panel
eDNA_dmm_structure(fit)
```

```{r struct_faceted}
# Faceted by true community, sorted by depth
# This lets you directly compare model assignments to the ground truth
p_struct <- eDNA_dmm_structure(
  fit,
  metadata  = sim$covariates,
  facet_var = "TrueCommunity",
  sort_var  = "Depth"
)
p_struct
```

With well-separated communities (which our simulation should produce), you
expect each panel to be dominated by one color. Mixed bars indicate either
genuinely ambiguous samples or label switching (communities are recovered but
numbered differently than the true labels).

```{r struct_custom_colors}
# Custom colors
eDNA_dmm_structure(
  fit,
  metadata         = sim$covariates,
  facet_var        = "TrueCommunity",
  community_colors = c(
    "Community 1" = "#E63946",
    "Community 2" = "#457B9D",
    "Community 3" = "#2A9D8F",
    "Community 4" = "#E9C46A"
  )
)
```

```{r struct_no_facet}
# No faceting — useful for real data where you don't know the true structure
# Sort by a covariate to reveal gradients
eDNA_dmm_structure(
  fit,
  metadata  = sim$covariates,
  sort_var  = "Depth"
)
```

```{r struct_save}
ggsave("structure_plot.png", p_struct, width = 10, height = 4, dpi = 300)
```

```{r struct_ggplot_extend}
# ggplot2 customization: change strip text size, add caption
p_struct +
  theme(strip.text  = element_text(size = 13, face = "bold")) +
  labs(caption = "Panels = true community | Sorted by depth within panel")
```

## Combining raw compositions with structure plot

Because both `plot_true_compositions()` and `eDNA_dmm_structure()` use the
same layout, they can be stacked directly with `cowplot` for a
before/after comparison:

```{r cowplot_combine, eval=FALSE}
library(cowplot)

p_raw    <- plot_true_compositions(counts, metadata = sim$covariates,
                                    facet_var = "TrueCommunity", sort_var = "Depth")
p_struct <- eDNA_dmm_structure(fit, metadata = sim$covariates,
                                facet_var = "TrueCommunity", sort_var = "Depth")

combined <- cowplot::plot_grid(
  p_raw, p_struct,
  ncol        = 1,
  labels      = c("A", "B"),
  label_size  = 16,
  rel_heights = c(1, 1)
)
cowplot::ggsave2("fig_compositions_vs_assignments.png",
                 combined, width = 10, height = 8, dpi = 300)
```

## NMDS ordination

```{r nmds}
nmds_result <- eDNA_dmm_nmds(fit)
nmds_result$plot

# Check stress value
message("NMDS stress: ", round(nmds_result$nmds$stress, 3))
# < 0.05 = excellent; 0.05-0.10 = good; 0.10-0.20 = adequate
```

```{r nmds_3d}
# 3D NMDS for lower stress; view different axis pairs
nmds_3d <- eDNA_dmm_nmds(fit, k = 3, nmds_axes = c(1, 3))
nmds_3d$plot
```

```{r nmds_jaccard}
# Jaccard distance instead of Bray-Curtis (presence/absence-weighted)
eDNA_dmm_nmds(fit, distance = "jaccard", show_ellipse = FALSE)$plot
```

```{r nmds_custom}
# Custom colors, no ellipses, larger points
eDNA_dmm_nmds(
  fit,
  show_ellipse     = FALSE,
  size_range       = c(2, 7),
  community_colors = c("Community 1" = "#E63946", "Community 2" = "#457B9D",
                       "Community 3" = "#2A9D8F", "Community 4" = "#E9C46A")
)$plot
```

```{r nmds_save}
ggsave("nmds.png", nmds_result$plot, width = 7, height = 6, dpi = 300)
```

## Beta coefficient plots

```{r beta_joint}
# Joint layout (default): all communities overlaid per covariate
beta_result <- eDNA_dmm_beta(fit)
beta_result$plot
```

```{r beta_separate}
# Separate layout: one row per community, one column per covariate
# Better for K > 3 where joint overlay gets crowded
eDNA_dmm_beta(fit, layout = "separate")$plot
```

```{r beta_table}
# Posterior summary table
beta_result$table
# Columns: community, covariate, mean, ci_5, ci_95,
#          prob_positive, prob_negative, ess, reliability
```

**Interpreting the beta plot:**

For our simulation, the true effects are:
- Community 1 (deep + offshore): high depth, high distance from shore
- Community 2 (surface + offshore): low depth, high distance from shore
- Community 3 (deep + inshore): high depth, low distance from shore
- Community 4 (surface + inshore, reference): this is the reference community

So relative to Community 4:
- Community 1 should have positive Depth beta and positive Distance_shore beta
- Community 2 should have negative Depth beta and positive Distance_shore beta
- Community 3 should have positive Depth beta and negative Distance_shore beta

Check whether the posteriors are pulled in these directions relative to the
Normal(0, 1) prior (grey).

```{r beta_specific}
# Plot only one covariate
eDNA_dmm_beta(fit, covariates_to_plot = "Depth")$plot

# Include the intercept
eDNA_dmm_beta(fit, show_intercept = TRUE)$plot

# Widen prior if you used a different prior SD in Stan
eDNA_dmm_beta(fit, beta_prior_sd = 2.5)$plot
```

```{r beta_save}
ggsave("beta_plot.png", beta_result$plot, width = 9, height = 5, dpi = 300)
```

---

# Verifying recovery of true parameters

One of the advantages of simulated data is that you can compare fitted values
to the truth. Here is a quick check.

## Community assignments

```{r verify_assignments}
# Compare MAP assignment to true community
comparison <- data.frame(
  sample_id     = fit$sample_info$sample_id,
  fitted_comm   = fit$sample_info$z_hat,
  true_comm     = paste0("Community ", sim$covariates$TrueCommunity),
  certainty     = round(fit$sample_info$assignment_certainty, 2)
)
print(comparison)

# Accuracy (ignoring label switching)
# Note: community labels may be permuted — Community 1 in fit ≠ Community 1 in truth
# A correct fit will show a 1:1 mapping between fitted and true, just possibly reordered
table(comparison$fitted_comm, comparison$true_comm)
```

## Alpha recovery

```{r verify_alpha}
# Simulated truth: alpha = 3 (set internally in simulate_eDNA_survey)
message("True alpha: 3.0")
message("Fitted alpha: ", round(fit$alpha_mean, 2))

# Full posterior
alpha_draws <- rstan::extract(fit$stan_fit, pars = "alpha")$alpha
hist(alpha_draws, breaks = 40, main = "Posterior for alpha",
     xlab = "alpha", col = "steelblue")
abline(v = 3.0, col = "red", lwd = 2, lty = 2)
legend("topright", legend = "True value", col = "red", lty = 2, lwd = 2)
```

---

# Working with real data

The workflow for real eDNA metabarcoding data is identical — you just skip
the simulation steps.

```{r real_data_workflow, eval=FALSE}
# 1. Load your count matrix (samples x taxa)
counts <- read.csv("my_count_table.csv", row.names = 1) |> as.matrix()

# 2. Load your metadata / covariates
metadata <- read.csv("my_metadata.csv")

# 3. Align them — critical step
counts   <- counts[order(rownames(counts)), ]
metadata <- metadata[order(metadata$sample_id), ]
stopifnot(all(rownames(counts) == metadata$sample_id))

# 4. Select covariates
cov_matrix <- metadata[, c("depth", "latitude")]

# 5. Select K using LOO
loo_result <- eDNA_loo(counts, cov_matrix, K_range = 2:5)
loo_result$plot

# 6. Fit the selected K
fit <- eDNA_dmm(counts, cov_matrix, K = 3)  # replace 3 with your chosen K

# 7. Visualize
eDNA_dmm_structure(fit, metadata = metadata, facet_var = "year",
                   sort_var = "latitude")
eDNA_dmm_nmds(fit)$plot
eDNA_dmm_beta(fit)$plot
```

---

# Troubleshooting

## Divergent transitions

```
Warning: There were X divergent transitions after warmup.
```

This means the sampler is struggling with the posterior geometry. Try:

1. `adapt_delta = 0.99` (increases step-size adaptation)
2. `max_treedepth = 15`
3. Try a lower K — more communities = harder geometry
4. Check for all-zero rows/columns in your count matrix

## Low ESS for beta parameters

ESS < 100 for beta coefficients usually means the covariate has little effect
on community membership. Look at the beta plot: if the posterior is nearly
identical to the prior (grey), the data do not support that covariate effect.
This is a valid scientific result, not a failure.

## Very long runtime

Each iteration requires N × K evaluations of the DirichletMultinomial likelihood,
which is expensive for large N and K. Practical tips:

- Start with `iter = 2000, warmup = 1000` for exploratory fits
- Use `iter = 4000+` for final inference
- Large S (many taxa) is less costly than large N or K
- Filter to top taxa by total reads to reduce S without losing major signal

## The model took too long and I killed it

Any partial `.rds` file written by `saveRDS()` will be incomplete. Restart
the fit with a fresh call to `eDNA_dmm()`. The compiled Stan model is cached,
so subsequent runs skip compilation.

---

# Session information

```{r session}
sessionInfo()
```
